---
title: "Introduction to the Tidyverse"
author: "Camila Bezerra"
date: "01/08/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo = F}
library(tidyverse)
library(dplyr)
library(gapminder)
library(ggplot2)
```

# Introduction to Cleaning Data in R

* Exploring raw data
* Removing or replacing missing data
* Preparing data for analysis

# Exploring raw data

* Understand the structure 
* Look at your data
* Visualize your data

# Understanding the structure of your data

```{r, eval = F}

#Load the lunch data
lunch <- read.csv("datasets/lunch_clean.csv")

#View its class
class(lunch)

#View its dimensions
dim(lunch)

#Look at column names
names(lunch)

#Look at the structure
str(lunch)

# Load dplyr
library(dplyr)

#View structure of lunch, the dplyr way
glimpse(lunch)

#View a summary
summary(lunch)


```


## Understanding the structure of your data

* class() - Class of data object
* dim() - Dimensions of data
* names() - Column names
* str() - Preview of data with helpful details
* glimpse() - Better vision of str() from dplyr
* summary() - Summary of data

## Exploring raw data (part2)

* Looking at your data

```{r, eval= F}

# View the top
head(lunch)

# View the bottom
tail(lunch)

# View histogram
hist(lunch$perc_free_red)

# View plot of two variables
plot(lunch$year, lunch$perc_free_red)

```

# Tidying data

## Introduction to tidy data

* Observations as rows
* Variables as columns
* One type of observational unit per table

## Principle of tidy data

* Each observation forms a row, each variable forms a column, and each type of observational unit forms a table.
* Each value belongs to a variable and an observation
* A dataset is a collection fo values

## Introduction to tidyr

* Gather columns into key-value pairs

```{r, eval=F}
# Gather the columns of wide_df
gather(wide_df, my_key, my_val, -col)

# Spread the key value pairs of long_df
spread(lon_df, my_key, my_val)

# Separate year_mo into two columns
separate(treatments, year_mo, c("year", "month"))

unite(treatments, year_mo, year, month)
```

## Summary of key tidyr functions

* gather() - Gather columns into key-value pairs
* spread() - Spread key-value pairs into columns
* separate() - Separate one column into multiple
* unite() - Unite multiple coumns into one

## Introduction to tidyr(part2)

* Column headers are values, not variable names
* Variables are stored in both rows and columns
* Multiple variables are stored in one column
* A single observational unit is stored in multiple tables
* Multiple types of observational units are stored in the same table

## Type conversions

Types of variables in R

* character:
* numeric: 23.44, 120, NaN, Inf
* integer: 4L, 1123L
* factor:
* logical:

Type conversions

* as.character()
* as.numeric()
* as.integer()
* as.factor()
* as.logical()

lubridate package

* coerce strings to date

```{r, eval=F}
library(lubridate)

ymd("2015-08-25")
ymd("2015 August 25")
mdy("August 25, 2015") 
hms("13:33:09")
ymd_hms("2015/08/25 13.33.09")
```

## String manipulation 

* stringr package

Key functions in stringr for cleaning data


```{r}
# str_trim() - Trim leading and trailing white space
str_trim("this is a test")

#str_pad() - Pad with additional characters
str_pad("24493", width = 7, side = "left", pad = "0")

# str_detect() - Search for string in vector

##Create character vector of names
friends <- c("Sarah", "Tom", "Alice")

##Search for string in vector
str_detect(friends, "Alice")


# str_replace() - Find and replace a pattern
str_replace(friends, "Alice", "David")

```


* Other helpful functions in base R

```{r}
# Make all lowercase
tolower("I AM TALKING LOUDLY")

#Make all uppercase
toupper("I am whispering")

```


## Missing and special values

* Finding missing values

```{r}
# Create small dataset

df <- data.frame(A = c(1, NA, 8, NA),
                 B = c(3, NA, 88, 23),
                 C = c(2, 45, 3, 1))

#Check for NAs
is.na(df)

#Are there any NAs?
any(is.na(df))

# Count number of NAs
sum(is.na(df))

# ou

# Use summary() to find NAs
summary(df)
```

* Dealing with missing values

```{r}
#Find rows with no missing values
complete.cases(df)

#Subset data, keeping only complete cases
df[complete.cases(df), ]

#Another way to remove rows with NAs
na.omit(df)
```


* Outliers and obvious errors


## Time to put it all together

* Get a feel for the data

```{r,eval=F}
# Verify that weather is a data.frame
class(weather)

# Check the dimensions
dim(weather)

# View the column names
names(weather)
```


* Summarize the data
```{r, eval = F}
str(weather)

# Load dplyr package
library(dplyr)

# Look at the structure using dplyr's glimpse()
glimpse(weather)

# View a summary of the data
summary(weather)

```

* Take a closer look

```{r, eval = F}
# View first 6 rows
head(weather)

# View first 15 rows
head(weather, n =15)

# View the last 6 rows
tail(weather)

# View the last 10 rows
tail(weather, n=10)
```


* Column names are values

```{r, eval = F}
# Load the tidyr package
library(tidyr)

# Gather the columns
weather2 <- gather(weather, day ,value, X1:X31, na.rm = TRUE)

# View the head
head(weather2)
```

* Values are variable names

```{r, eval = F}
# First remove column of row names
without_x <- weather2[, -1]

# Spread the data
weather3 <- spread(without_x, measure, value)

# View the head
head(weather3)
```

* Clean up dates

```{r, eval = F}
# Load the stringr and lubridate packages
library(stringr)
library(lubridate)


# Remove X's from day column
weather3$day <- str_replace(weather3$day, "X", "")

# Unite the year, month, and day columns
weather4 <- unite(weather3, date, year, month, day, sep = "-")

# Convert date column to proper date format using lubridates's ymd()
weather4$date <- ymd(weather4$date)

# Rearrange columns using dplyr's select()
weather5 <- select(weather4, date, Events, CloudCover:WindDirDegrees)

# View the head of weather5
head(weather5)
```

* A closer look at column types
```{r, eval = F}
# View the structure of weather5
str(weather5)

# Examine the first 20 rows of weather5. Are most of the characters numeric?
head(weather5, n= 20)

# See what happens if we try to convert PrecipitationIn to numeric
as.numeric(weather5$PrecipitationIn)

```

*Column type conversions

```{r, eval = F}
# Replace "T" with "0" (T = trace)
weather5$PrecipitationIn <- replace(weather5$PrecipitationIn, "T", "0")

# Convert characters to numerics
weather6 <- mutate_at(weather5, vars(CloudCover:WindDirDegrees), funs(as.numeric))

# Look at result
str(weather6)
```

## Missing, extreme, and unexpected values

* Find missing values

```{r, eval = F}
# Count missing values
sum(is.na(weather6))

# Find missing values
summary(weather6)

# Find indices of NAs in Max.Gust.SpeedMPH
ind <- which(is.na(weather6$Max.Gust.SpeedMPH))

# Look at the full rows for records missing Max.Gust.SpeedMPH
weather6[ind, ]
```


* An obvious error

```{r, eval = F}

# Review distributions for all variables
summary(weather6)

# Find row with Max.Humidity of 1000
ind <- which(weather6$Max.Humidity == 1000)

# Look at the data for that day
weather6[ind, ]

# Change 1000 to 100
weather6$Max.Humidity[ind] <- 100
```

* Another obvious error

```{r, eval = F}

# Look at summary of Mean.VisibilityMiles
summary(weather6$Mean.VisibilityMiles)


# Get index of row with -1 value
ind <- which(weather6$Mean.VisibilityMiles == -1)

# Look at full row
weather6[ind, ]

# Set Mean.VisibilityMiles to the appropriate value
weather6$Mean.VisibilityMiles[ind] <- 10
```

* Check other extreme values

```{r, eval=F}
# Review summary of full data once more
summary(weather6)

# Look at histogram for MeanDew.PointF
hist(weather6$MeanDew.PointF)

# Look at histogram for Min.TemperatureF
hist(weather6$Min.TemperatureF)

# Compare to histogram for Mean.TemperatureF
hist(weather6$Mean.TemperatureF)
```

* Finishing touches

```{r, eval = F}
# Clean up column names
names(weather6) <- new_colnames

# Replace empty cells in events column
weather6$events[weather6$events == ""] <-"None"
    
# Print the first 6 rows of weather6
head(weather6)
```

## Summary of your accomplishments

* Inspect the data
* Tidied the data
* Improved date representations
* Dealt with incorrect variable codings
* Found and dealt with missing values
* Identified and corrected errors
* Visualized the result