---
title: "Cluster Analysis in R"
author: "Camila Bezerra"
date: "09/09/2019"
output:
  html_document:
    theme: united
    highlight: monochrome
---

```{r, warning=F}
library(ggplot2)
library(tidyverse)
library(dplyr)
```


# 1. Calculating distance between observations

## Calculate & plot the distance between two players

```{r}
# Plot the positions of the players
ggplot(two_players, aes(x = x, y = y)) + 
  geom_point() +
  # Assuming a 40x60 field
  lims(x = c(-30,30), y = c(-20, 20))

# Split the players data frame into two observations
player1 <- two_players[1, ]
player2 <- two_players[2, ]

# Calculate and print their distance using the Euclidean Distance formula
player_distance <- sqrt( (player1$x - player2$x)^2 + (player1$y - player2$y)^2 )
player_distance
```


## Calculating distance between categorical variables

```{r}
# Dummify the Survey Data
dummy_survey <- dummy.data.frame(job_survey)

# Calculate the Distance
dist_survey <- dist(dummy_survey, method = 'binary')

# Print the Original Data
job_survey
```

# 2. Hierarchical clustering

## Comparing more than two observations

### Linkage criteria

* Complete Linkage: maximum distance between two sets
* Single linkage: minimum distance between two sets
* Average Linkage: average distance between two sets

```{r}
# Extract the pair distances
distance_1_2 <- dist_players[1]
distance_1_3 <- dist_players[2]
distance_2_3 <- dist_players[3]

# Calculate the complete distance between group 1-2 and 3
complete <- max(c(distance_1_3, distance_2_3))
complete

# Calculate the single distance between group 1-2 and 3
single <- min(c(distance_1_3, distance_2_3))
single

# Calculate the average distance between group 1-2 and 3
average <- mean(c(distance_1_3, distance_2_3))
average
```

## Capturing K clusters

### Extracting K Clusters

```{r}
cluster_assignments <- cutree(hc_players, k = 2)

palyers_clustered <- mutate(players, cluster = cluster_assignments)
```


## Assign cluster membership

```{r}
# Calculate the Distance
dist_players <- dist(lineup, method = 'euclidean')

# Perform the hierarchical clustering using the complete linkage
hc_players <- hclust(dist_players, method="complete")

# Calculate the assignment vector with a k of 2
clusters_k2 <- cutree(hc_players, k = 2)

# Create a new data frame storing these results
lineup_k2_complete <- mutate(lineup, cluster = clusters_k2)
```


## Exploring the clusters

```{r}
# Count the cluster assignments
count(lineup_k2_complete, cluster)

# Plot the positions of the players and color them using their cluster
ggplot(lineup_k2_complete, aes(x = x, y = y, color = factor(cluster))) +
  geom_point()
```


## Comparing average, simple & complete linkage

```{r}
# Prepare the Distance Matrix
dist_players <- dist(lineup)

# Generate hclust for complete, single & average linkage methods
hc_complete <- hclust(dist_players, method = "complete")
hc_single <- hclust(dist_players, method = "single")
hc_average <- hclust(dist_players, method = "average")

# Plot & Label the 3 Dendrograms Side-by-Side
# Hint: To see these Side-by-Side run the 4 lines together as one command
par(mfrow = c(1,3))
plot(hc_complete, main = 'Complete Linkage')
plot(hc_single, main = 'Single Linkage')
plot(hc_average, main = 'Average Linkage')
```


## Cutting the tree

### Coloring the Dendogram - Height

```{r}
library(dendextend)
dend_players <- as.dendrogram(hc_players)
dend_colored <- color_branches(dend_players, h =10)
plot(dend_colored)
```

### cutree() using height

```{r}
cluster_assignments <- cutree(hc_players, h = 15)
```


## Clusters based on height

```{r}
library(dendextend)
dist_players <- dist(lineup, method = 'euclidean')
hc_players <- hclust(dist_players, method = "complete")

# Create a dendrogram object from the hclust variable
dend_players <- as.dendrogram(hc_players)

# Plot the dendrogram
plot(dend_players)

# Color branches by cluster formed from the cut at a height of 20 & plot
dend_20 <- color_branches(dend_players, h = 20)

# Plot the dendrogram with clusters colored below height 20
plot(dend_20)

# Color branches by cluster formed from the cut at a height of 40 & plot
dend_40 <- color_branches(dend_players, h = 40)

# Plot the dendrogram with clusters colored below height 40
plot(dend_40)
```

## Segment wholesale customers

```{r}
# Calculate Euclidean distance between customers
dist_customers <- dist(customers_spend, method = "euclidean")

# Generate a complete linkage analysis 
hc_customers <- hclust(dist_customers, method = "complete")

# Plot the dendrogram
plot(hc_customers)

# Create a cluster assignment vector at h = 15000
clust_customers <- cutree(hc_customers, h = 15000)

# Generate the segmented customers data frame
segment_customers <- mutate(customers_spend, cluster = clust_customers)
```

# 3. K-means clustering

### kmeans()

```{r}
model <- kmeans(lineup, centers = 2)
lineup_clustered <- mutate(lineup, cluster = model$cluster)
```

## K-means on a soccer field

```{r}
# Build a kmeans model
model_km2 <- kmeans(lineup, centers = 2)

# Extract the cluster assignment vector from the kmeans model
clust_km2 <- kmeans(lineup, centers = 2)

# Create a new data frame appending the cluster assignment
lineup_km2 <- mutate(lineup, cluster = model_km2$cluster)

# Plot the positions of the players and color them using their cluster
ggplot(lineup_km2, aes(x = x, y = y, color = factor(cluster))) +
  geom_point()
```

## K-means on a soccer field(part2)

```{r}
# Build a kmeans model
model_km3 <- kmeans(lineup, centers = 3)

# Extract the cluster assignment vector from the kmeans model
clust_km3 <- kmeans(lineup, centers = 3)

# Create a new data frame appending the cluster assignment
lineup_km3 <- mutate(lineup, cluster = model_km3$cluster)

# Plot the positions of the players and color them using their cluster
ggplot(lineup_km3, aes(x = x, y = y, color = factor(cluster))) +
  geom_point()
```

## Evaluating different values of K by eye

```{r}
# Generating the Elbow plot

model <- kmeans(x = lineup, ccenters = 2)

# Generating the Elbow Plot
library(purrr)
tot_withinss <- map_dbl(1:10, function(k)){
  model <- kmeans(x = lineup, centers = k)
  model$tot.withinss
}
```

## Many k's many models

```{r}
library(purrr)

# Use map_dbl to run many models with varying value of k (centers)
tot_withinss <- map_dbl(1:10,  function(k){
  model <- kmeans(x = lineup, centers = k)
  model$tot.withinss
})

# Generate a data frame containing both k and tot_withinss
elbow_df <- data.frame(
  k = 1:10 ,
  tot_withinss = tot_withinss
)
```

## Elbow (Scree) plot
```{r}
# Use map_dbl to run many models with varying value of k (centers)
tot_withinss <- map_dbl(1:10,  function(k){
  model <- kmeans(x = lineup, centers = k)
  model$tot.withinss
})

# Generate a data frame containing both k and tot_withinss
elbow_df <- data.frame(
  k = 1:10,
  tot_withinss = tot_withinss
)

# Plot the elbow plot
ggplot(elbow_df, aes(x = k, y = tot_withinss)) +
  geom_line() +
  scale_x_continuous(breaks = 1:10)
```

## Silhouette analysis: Observation level performance

```{r}

library(cluster)

pam_k3 <- pam(lineup, k = 3)

# Silhouette Plot
sil_plot <- silhouette(pam_k3)
plot(sil_plot)

# Average Silhouette Width
pam_k3$silinginfo$avg.width
```

### Highest Average Silhouette Width

```{r}
library(purrr)

sil_width <- map_dbl(2:10, function(k){
  model <- pam(x = lineup, k = k)
  model$silinginfo$avg.width
})

sil_df <- data.frame(
  k = 2:10,
  stil_width = sil_width
)

print(sil_df)
```

## Silhouette analysis

```{r}
library(cluster)

# Generate a k-means model using the pam() function with a k = 2
pam_k2 <- pam(lineup, k = 2)

# Plot the silhouette visual for the pam_k2 model
plot(silhouette(pam_k2))

# Generate a k-means model using the pam() function with a k = 3
pam_k3 <-pam(lineup, k = 3)

# Plot the silhouette visual for the pam_k3 model
plot(silhouette(pam_k3))
```

## Revisiting wholesale data: "Best" k

```{r}
# Use map_dbl to run many models with varying value of k
sil_width <- map_dbl(2:10,  function(k){
  model <- pam(x = customers_spend, k = k)
  model$silinfo$avg.width
})

# Generate a data frame containing both k and sil_width
sil_df <- data.frame(
  k = 2:10,
  sil_width = sil_width
)

# Plot the relationship between k and sil_width
ggplot(sil_df, aes(x = k, y = sil_width)) +
  geom_line() +
  scale_x_continuous(breaks = 2:10)
```

## Revisiting wholesale data: Exploration

```{r}
set.seed(42)

# Build a k-means model for the customers_spend with a k of 2
model_customers <- kmeans(customers_spend, centers = 2)

# Extract the vector of cluster assignments from the model
clust_customers <- kmeans(customers_spend, centers = 3)

# Build the segment_customers data frame
segment_customers <- mutate(customers_spend, cluster = model_customers$cluster)

# Calculate the size of each cluster
count(segment_customers, cluster)

# Calculate the mean for each category
segment_customers %>% 
  group_by(cluster) %>% 
  summarise_all(funs(mean(.))klç)
```

# 4. Case Study: National Occupational mean wage

###################REVIEW THIS##
```{r}
# Calculate Euclidean distance between the occupations
dist_oes <- dist(oes, method = "euclidean")

# Generate an average linkage analysis 
hc_oes <- hclust(oes, method = "average")

# Create a dendrogram object from the hclust variable
dend_oes <- as.dendrogram(hc_oes)

# Plot the dendrogram
plot(dend_oes)
```

## Hierarchical clustering: Occupation trees

```{r}
# Calculate Euclidean distance between the occupations
dist_oes <- dist(oes, method = "euclidean")

# Generate an average linkage analysis 
hc_oes <- hclust(dist_oes, method = "average")

# Create a dendrogram object from the hclust variable
dend_oes <- as.dendrogram(hc_oes)

# Plot the dendrogram
plot(dend_oes)

dend_colored <- (color_branches(dend_oes, h = 100000)) 

plot(dend_colored)
```

## Hierarchical clustering: Preparing for exploration

```{r}
dist_oes <- dist(oes, method = 'euclidean')
hc_oes <- hclust(dist_oes, method = 'average')

library(tibble)
library(tidyr)

# Use rownames_to_column to move the rownames into a column of the data frame
df_oes <- rownames_to_column(as.data.frame(oes), var = 'occupation')

# Create a cluster assignment vector at h = 100,000
cut_oes <- cutree(hc_oes, h = 100000)

# Generate the segmented the oes data frame
clust_oes <- mutate(df_oes, cluster = cut_oes)

# Create a tidy data frame by gathering the year and values into two columns
gathered_oes <- gather(data = clust_oes, 
                       key = year, 
                       value = mean_salary, 
                       -occupation, -cluster)
```

## Hierarchical clustering: Plotting occupational clusters

```{r}
# View the clustering assignments by sorting the cluster assignment vector
sort(cut_oes)

# Plot the relationship between mean_salary and year and color the lines by the assigned cluster
ggplot(gathered_oes, aes(x = year, y = mean_salary, color = factor(cluster))) + 
    geom_line(aes(group = occupation))
```

## K-means: Elbow analysis

```{r}
Use map_dbl to run many models
```

## K-means: Average Silhouette Widths

```{r}
# Use map_dbl to run many models with varying value of k
sil_width <- map_dbl(2:10,  function(k){
  model <- pam(oes, k = k)
  model$silinfo$avg.width
})

# Generate a data frame containing both k and sil_width
sil_df <- data.frame(
  k = 2:10,
  sil_width = sil_width
)

# Plot the relationship between k and sil_width
ggplot(sil_df, aes(x = k, y = sil_width)) +
  geom_line() +
  scale_x_continuous(breaks = 2:10)
```

