---
title: "Communicating with Data in the Tidyverse"
author: "Camila Bezerra"
date: "28/08/2019"
output:
  html_document:
    theme: united
    highlight: monochrome
---

```{r, echo=FALSE}
knitr::opts_chunk$set(eval = F)
```

```{r, warning=F}
library(ggplot2)
library(mtcars)
```

# 1. Custom ggplot2 themes

## Join the two data sets together

```{r, eval = F}
# Join both data frames
ilo_data <- ilo_hourly_compensation %>%
  inner_join(ilo_working_hours, by = c("country", "year"))

# Count the resulting rows
ilo_data  %>% 
    count()

# Examine ilo_data
ilo_data
```


## Change variable types

```{r, eval = F}
# Turn year and country into a factor
ilo_data_corrected <- ilo_data %>%
  mutate(year = as.factor(as.numeric(year)),
        country = as.factor(country))
```

## Filter the data for plotting

```{r, eval = F}
# Examine the European countries vector
european_countries

# Only retain European countries
ilo_data <- ilo_data %>% 
  filter(country %in% c (european_countries))
```

## Some summary statistics

```{r, eval = F}
# Examine the structure of ilo_data
str(ilo_data)

# Group and summarize the data
ilo_data %>%
  group_by(year) %>%
  summarize(mean_hourly_compensation = mean(hourly_compensation),
            mean_working_hours = mean(working_hours))

```

## A basic scatter plot

```{r, eval = F}
# Filter for 2006
plot_data <- ilo_data %>%
  filter(year == 2006)
  
# Create the scatter plot
ggplot(plot_data) +
  geom_point(aes(x = working_hours, y = hourly_compensation))
```

## Add labels to the plot

```{r, eval = F}
#Create the plot
ggplot(plot_data) +
  geom_point(aes(x = working_hours, y = hourly_compensation)) +
  # Add labels
  labs(x = "Working hours per week",
       y = "Hourly compensation",
       title = "The more people work, the less compensation thet seem to receive",
       subtitle = "Working hours and hourly compensation in European countries, 2006",
       caption = "Data source: ILO, 2017")
```

## Apply a default theme

```{r, eval = F}
# Save your current plot into a variable: ilo_plot
ilo_plot <- ggplot(plot_data) +
  geom_point(aes(x = working_hours, y = hourly_compensation)) +
  labs(
    x = "Working hours per week",
    y = "Hourly compensation",
    title = "The more people work, the less compensation they seem to receive",
    subtitle = "Working hours and hourly compensation in European countries, 2006",
    caption = "Data source: ILO, 2017"
  )

# Try out theme_minimal
ilo_plot +
  theme_minimal()

# Try out any other possible theme function
ilo_plot +
  theme_bw()
```

## Change the appearance of titles

```{r, eval = F}
ilo_plot <- ilo_plot +
  theme_minimal() +
  # Customize the "minimal" theme with another custom "theme" call
  theme(
    text = element_text(family = "Bookman"),
    title = element_text(color = "gray25"),
    plot.caption = element_text(color = "gray30"),
    plot.subtitle = element_text(size = 12)
  )

# Render the plot object
ilo_plot
```

## Alter background color and add margins

Change the overall background color of your plot to "gray95".
This time, another function than element_text is needed – one for rectangular plot elements. Rewatch the video to know which.
Look up the correct sequence of margin specifications in the documentation by typing ?theme.
Using these specifications, add margins to the plot: 5mm top and bottom, 10mm to the left and the right.

```{r, eval = F}
ilo_plot +
  # "theme" calls can be stacked upon each other, so this is already the third call of "theme"
  theme(
    plot.background = element_rect(fill = "gray95"),
    plot.margin = unit(c(5, 10, 5, 10), units = "mm")
  )
```

# 2. Creating a custom and unique visualization

## Visualizing aspects of data with facets

* The facet_grid() function

    * how to facet by columns

```{r, eval = F}
+ 
facet_grid(. ~ year)

+
facet_wrap(facets = ~ year)
```

   *how to facet by rows
   
```{r, eval = F}
+
facet_grid(year ~.)
```

* Defining your own theme function

```{r, eval = F}
theme_green <- function{
  theme(
    plot.background = 
      element_rect(fill = "green"),
    panel.background =
      element_rect(fill =
                     "lightgreen")
  )
}
```

## Add facets to the plot

```{r, eval = F}
# Again, you save the plot object into a variable so you can save typing later on
ilo_plot <- ggplot(ilo_data, aes(x = working_hours, y = hourly_compensation)) +
  geom_point() +
   labs(
    x = "Working hours per week",
    y = "Hourly compensation",
    title = "The more people work, the less compensation they seem to receive",
    subtitle = "Working hours and hourly compensation in European countries, 2006",
    caption = "Data source: ILO, 2017"
  ) +
  # Add facets here
  facet_grid(.~ year)
```

## Define your own theme function

```{r, eval = F}
# For a starter, let's look at what you did before: adding various theme calls to your plot object
ilo_plot +
  theme_minimal() +
  theme(
    text = element_text(family = "Bookman", color = "gray25"),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(color = "gray30"),
    plot.background = element_rect(fill = "gray95"),
    plot.margin = unit(c(5, 10, 5, 10), units = "mm")
  )
      
# Define your own theme function below
theme_ilo <- function() {
    theme_minimal() +
  theme(
    text = element_text(family = "Bookman", color = "gray25"),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(color = "gray30"),
    plot.background = element_rect(fill = "gray95"),
    plot.margin = unit(c(5, 10, 5, 10), units = "mm")
  )
}
```

## Apply the new theme function to the plot

```{r, eval = F}
# Apply your theme function (dont't forget to call it with parentheses!)
ilo_plot <- ilo_plot +
  theme_ilo()

# Examine ilo_plot
ilo_plot

ilo_plot +
  # Add another theme call
  theme(
    # Change the background fill and color
    strip.background = element_rect(fill = "gray60", colour = "gray95"),
    # Change the color of the text
    strip.text = element_text(color = "white")
  )
```

## A custom plot to emphasize change

geom_path() connects the observations in the order in which they appear in the data.

```{r, eval = F}
ggplot() +
  geom_path(aes(x = numeric_variable, y = numeric_variable))

ggplot() +
 geom_path(aes(x = numeric_variable, y = factor_variable))

ggplot() +
  geom_path(aes(x = numeric_variable, y = factor_variable),
            arrow = arrow (----))
  
```

## A basic dot plot

```{r, eval = F}
# Create the dot plot
ggplot()+
geom_path(aes( x = working_hours, y = country)
```

## Add arrows to the lines in the plot

Use the arrow argument and an arrow() function call to add arrows to each line.
For the arrows, specify a length of 1.5 "mm" (with a unit() call) and a type of "closed".

```{r, eval = F}
ggplot(ilo_data) +
  geom_path(aes(x = 
                  working_hours, y = country),
            # Add an arrow to each path
            arrow = arrow (length = unit(1.5, "mm"), type = "closed"))
```

## Add some labels to each country

Label both points on the line with their respective working hours value. Use the geom_text() function for this, with the x- and y-aesthetic identical to the one in the geom_path() call.
Specify the working_hours variable for the label argument. Round working_hours to one digit after the decimal point.

```{r, eval = F}
ggplot(ilo_data) +
  geom_path(aes(x = working_hours, y = country),
            arrow = arrow(length = unit(1.5, "mm"), type = "closed")) +
  # Add a geom_text() geometry
  geom_text(
          aes(x = working_hours,
              y = country,
              label = round(working_hours, 1))
        )
```

now the x-axis is obsolete and can be removed in the remainder of this chapter. The labels are still kind of misplaced, though… Let's fix this!

## Polishing the dot plot

* Factor levels

    * The order of factor levels determine the order of appearence in ggplot2.

* The fct_reorder function

```{r, eval = F}
ilo_data <- ilo_data %>% 
  mutate(country = fct_reorder(cuntry, working_hours, mean))
```


## Reordering elements in the plot

```{r, eval = F}
library(forcats)

# Reorder country factor levels
ilo_data <- ilo_data %>%
  # Arrange data frame
  arrange(year) %>%
  # Reorder countries by working hours in 2006
  mutate(country = fct_reorder(country,
                               working_hours,
                               last))

# Plot again
ggplot(ilo_data) +
  geom_path(aes(x = working_hours, y = country),
            arrow = arrow(length = unit(1.5, "mm"), type = "closed")) +
    geom_text(
          aes(x = working_hours,
              y = country,
              label = round(working_hours, 1))
          )
```

## Reordering elements in the plot

```{r, eval = F}
library(forcats)

# Reorder country factor levels
ilo_data <- ilo_data %>%
  # Arrange data frame
  arrange(year) %>%
  # Reorder countries by working hours in 2006
  mutate(country = fct_reorder(country,
                               working_hours,
                               last))

# Plot again
ggplot(ilo_data) +
  geom_path(aes(x = working_hours, y = country),
            arrow = arrow(length = unit(1.5, "mm"), type = "closed")) +
    geom_text(
          aes(x = working_hours,
              y = country,
              label = round(working_hours, 1))
          )

```


## Correct ugly label positions

```{r, eval = F}
# Save plot into an object for reuse
ilo_dot_plot <- ggplot(ilo_data) +
  geom_path(aes(x = working_hours, y = country),
            arrow = arrow(length = unit(1.5, "mm"), type = "closed")) +
    # Specify the hjust aesthetic with a conditional value
    geom_text(
          aes(x = working_hours,
              y = country,
              label = round(working_hours, 1),
              hjust = ifelse(year == "2006", 1.4, -0.4)
            ),
          # Change the appearance of the text
          size = 3,
          family = "Bookman",
          color = "gray25"
          )

ilo_dot_plot
```


## Finalizing the plot for different audiences and devices

* coord_cartesian vs. xlim / ylim 

```{r, eval = F}
ggplot_object +
  coord_cartesian(xlim = c(0, 100), ylim = c(10,20))

# or
ggplot_object +
  xlim(0, 100) +
  ylim(10, 20)
```


## Change the viewport so labels don't overlap with plot border

```{r, eval = F}
# Reuse ilo_dot_plot
ilo_dot_plot <- ilo_dot_plot +
  #Add labels to the plot
  labs(
    x = "Working hours per week",
    y = "Country",
    title = "People work less in 2006 compared to 1996",
    susubtitle = "Working hours in European countries, development since 1996",
    caption = "Data source: ILO, 2017"
  ) +
  # Apply your theme
  theme_ilo() +
  # Change the viewport
  coord_cartesian(xlim = c(25, 41))
  ))

#View the plot
ilo_dot_plot
```

## Optimizing the plot for mobile devices

```{r, eval = F}
# Compute temporary data set for optimal label placement
# Compute temporary data set for optimal label placement
median_working_hours <- ilo_data %>%
  group_by(country) %>%
  summarize(median_working_hours_per_country = median(working_hours)) %>%
  ungroup()

# Have a look at the structure of this data set
str(median_working_hours)

ilo_dot_plot +
  # Add label for country
  geom_text(data = median_working_hours,
            aes(y = country,
                x = median_working_hours_per_country,
                label = country),
            vjust = 2,
            family = "Bookman",
            color = "gray25") +
  # Remove axes and grids
  theme(
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank(),
    # Also, let's reduce the font size of the subtitle
    plot.subtitle = element_text(size = 9)
  )
```

# 3. Introduction to RMarkdown

## Change formatting of text snippets

* Turn "data sets" in the first sentence into a link to http://www.ilo.org/global/statistics-and-databases/lang--en/index.htm

The **International Labour Organization** (ILO) has many [data sets] (link) on working conditions.

## Images in RMarkdown files

### Adjusting figure options in RMarkdown

* fig.height = numeric, inches: the height of the image in in inches.
* fig.width = numeric, inches: the width of the image in inches
* fig.align = "left", "right", "center": the alignment of the image in the report.

```{r, eval = F, fig.height= 6}

ggplot2 code

```

### External images in RMarkdown reports

![Corcovado](https://images.app.goo.gl/EEfvgFCXkMv15AQ36)


# 4. Customizing your RMarkdown report

## Add a table of contents

```{r, eval = F}
---
title: "The reduction in weekly working hours in Europe" 
subtitle: "Looking at the development between 1996 and 2006"
author: "Insert your name here"
output: 
  html_document:
    theme: cosmo
    highlight: monochrome
    toc: true
    toc_float: false
    toc_depth: 4
---

```

## More YAML hacks

Hide all the code chunks by default by specifying the appropriate key-value-pair in the YAML header.
Use code_folding: ... for this according to the instructions above.
Try out another setting: Add numbers to all sections with the number_sections option. Remember: All these settings hierarchically belong to the html_document key.

```{r, eval = F}
---
title: "The reduction in weekly working hours in Europe" 
subtitle: "Looking at the development between 1996 and 2006"
author: "Insert your name here"
output: 
  html_document:
    theme: cosmo
    highlight: monochrome
    toc: true
    toc_float: false
    toc_depth: 4
    code_folding: hide
    number_sections: true
---

```


## CSS selectors

What CSS selector would be needed to match all table header cells (th) that are direct children of a table with the class beautiful?

```{r, eval = F}

table.beautiful > th
```


## Change style attributes of text elements

On line 17, change the font-family of all the text in your report (including headers, except R code chunks) to "Bookman", serif.
On line 21, change the color of the body text to a slight grey (#333333).
On line 24, change the color of all links to red.
On line 27, reduce the font of the R code elements, which are wrapped in pre HTML tags, to 10px.

<style>
body, h1, h2, h3, h4 {
    font-family: "Bookman", serif;
}

body {
    color: #333333;
}
a, a:hover {
    color: red;
}
pre {
    font-size: 10px;
}
</style>

## Reference the style sheet

```{r, eval = F}
---
title: "The reduction in weekly working hours in Europe" 
subtitle: "Looking at the development between 1996 and 2006"
author: "Insert your name here"
output: 
  html_document:
    css: styles.css
    theme: cosmo
    highlight: monochrome
    toc: true
    toc_float: false
    toc_depth: 4
    code_folding: hide
---
```



## Beautify a table with kable

Use the knitr::kable() function call to render the table in the "Data" section with the kable engine (in the chunk that starts on line 40).

```{r generating_summary_statistics, echo = TRUE, eval=F}
# Some summary statistics
ilo_data %>%
  group_by(year) %>%
  summarize(mean_hourly_compensation = mean(hourly_compensation),
            mean_working_hours = mean(working_hours)) %>%
  # pipe the above data frame into the knitr::kable function
  knitr::kable()

```


## Repetition: CSS

What CSS rule would be needed to color the front in the header row to "red"?

```{r, eval = F}
th { color: red; }
```


