---
title: "Data Visualization with ggplot2 (Part 1)"
author: "Camila Bezerra"
date: "27/08/2019"
output: html_document
---

```{r, echo=FALSE}
knitr::opts_chunk$set(eval = F)
```

```{r, warning=F}
library(ggplot2)
library(mtcars)
```

# 1. Introduction

## Exploring ggplot2, part 2 

```{r}
# Change the command below so that cyl is treated as factor
## This time the x-axis does not containg variables like 5 or 7, only the values that are present in the dataset.
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_point()
```


## Understanding Variables

shape: only makes sense with categorical data, and disp is continuous.

## Exploring ggplot2, part 4

Use the + operator to add geom_point() and geom_smooth(). These just stack on each other! geom_smooth() will draw a smoothed line over the points.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
geom_point()+
geom_smooth()

```

## Undestanding the grammar, part 1
```{r, eval = F}
# Create the object containing the data and aes layers: dia_plot
dia_plot <- ggplot(diamonds, aes(x = carat, y = price))

# Add a geom layer with + and geom_point()
dia_plot + geom_point()

# Add the same geom layer, but with aes() inside
dia_plot + geom_point(aes(color = clarity))

```

## Understandig the grammar, part 2

To set a propoerty of a geom to a single value, pass it as an argument.

To give the property different values for each row of data, pass it as na aesthetic.

```{r, eval = F}
# 1 - The dia_plot object has been created for you
dia_plot <- ggplot(diamonds, aes(x = carat, y = price))

# 2 - Expand dia_plot by adding geom_point() with alpha set to 0.2
dia_plot <- dia_plot + geom_point(alpha = 0.2)

# 3 - Plot dia_plot with additional geom_smooth() with se set to FALSE
dia_plot + geom_smooth(se = F)

# 4 - Copy the command from above and add aes() with the correct mapping to geom_smooth()
dia_plot + geom_smooth(aes(col = clarity), se = F)
```

# 2. Data

## Objects and Layers

```{r, eval = F}
# Plot the correct variables of mtcars
plot(mtcars$wt, mtcars$mpg, col = mtcars$cyl)

# Change cyl inside mtcars to a factor
mtcars$fcyl <- as.factor(mtcars$cyl)

# Make the same plot as in the first instruction
plot(mtcars$wt, mtcars$mpg, col = mtcars$fcyl)
```

## base package and ggplot2, part2 - lm

```{r, eval = F}
# Use lm() to calculate a linear model and save it as carModel
carModel <- lm(mpg ~ wt, data = mtcars)

# Basic plot
mtcars$cyl <- as.factor(mtcars$cyl)
plot(mtcars$wt, mtcars$mpg, col = mtcars$cyl)

# Call abline() with carModel as first argument and set lty to 2
abline(carModel, lty = 2)

# Plot each subset efficiently with lapply
# You don't have to edit this code
plot(mtcars$wt, mtcars$mpg, col = mtcars$cyl)
lapply(mtcars$cyl, function(x) {
  abline(lm(mpg ~ wt, mtcars, subset = (cyl == x)), col = x)
  })

# This code will draw the legend of the plot
# You don't have to edit this code
legend(x = 5, y = 33, legend = levels(mtcars$cyl),
       col = 1:3, pch = 1, bty = "n")
```

## Variabels to visual, part 1

```{r, eval = F}
ggplot(iris.tidy, aes(x = Species, y = value, col = Part)) +
  geom_jitter() +
  facet_grid(. ~ Measure)
```

## Variables to visuals, part 1b

```{r, eval = F}
# Load the tidyr package
library(tidyr)

# Fill in the --- to produce to the correct iris.tidy dataset
iris.tidy <- iris %>% 
  gather(key, Value, -Species) %>% 
  separate(key, c("Part", "Measure"), "\\.")

```

## Variables to visuals, part 2

```{r, eval = F}
ggplot(iris.wide, aes(x = Length, y = Width, color = Part)) +
  geom_jitter() +
  facet_grid(. ~ Species)
```

## Variables to visuals, part 2b

gather() rearranges the data frame by specifying the columns that are categorical variables with a - notation. In this case, Species and Flower are categorical. Complete the command.
separate() splits up the new key column, which contains the former headers, according to .. The new column names "Part" and "Measure" are given in a character vector.
The last step is to use spread() to distribute the new Measure column and associated value column into two columns.

```{r, eval = F}
# Load the tidyr package
library(tidyr)

# Add column with unique ids (don't need to change)
iris$Flower <- 1:nrow(iris)

# Fill in the ___ to produce to the correct iris.wide dataset
iris.wide <- iris %>%
  gather(key, value, -Species, -Flower) %>%
  separate(key, c("Part", "Measure"), "\\.") %>%
  spread("Measure", value)
```

# 3. Aesthetics

## Visible Aesthetics

* x: X axis position
* y: Y axis position
* colour: Colour of dots
* fill: fill coulour
* size: diameter of points
* alpha: transparency
* linetype: Line dash pattern
* labels: Text on a plot or axes
* shape: shape

## All about aesthetics, part 1

```{r, eval = F}
# 1 - Map mpg to x and cyl to y
ggplot(mtcars, aes(mpg, cyl)) +
  geom_point()
  
# 2 - Reverse: Map cyl to x and mpg to y
ggplot(mtcars, aes(cyl, mpg)) +
  geom_point()

# 3 - Map wt to x, mpg to y and cyl to col
ggplot(mtcars, aes(wt, mpg, col = cyl)) +
  geom_point()

# 4 - Change shape and size of the points in the above plot
ggplot(mtcars, aes(wt, mpg, col = cyl)) +
  geom_point(shape = 1, size = 4)

```

## All about aesthetics, part 2

Shapely coding! Notice that mapping a categorical variable onto fill doesn't change the colors, although a legend is generated! This is because the default shape for points only has a color attribute and not a fill attribute! Use fill when you have another shape (such as a bar), or when using a point that does have a fill and a color attribute, such as shape = 21, which is a circle with an outline. Any time you use a solid color, make sure to use alpha blending to account for over plotting.

```{r, eval = F}
# am and cyl are factors, wt is numeric
class(mtcars$am)
class(mtcars$cyl)
class(mtcars$wt)

# From the previous exercise
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
  geom_point(shape = 1, size = 4)

# 1 - Map cyl to fill
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl)) +
  geom_point(shape = 1, size = 4)


# 2 - Change shape and alpha of the points in the above plot
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl)) +
  geom_point(shape = 21, size = 4, alpha = 0.6)


# 3 - Map am to col in the above plot
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl, col = am)) +
  geom_point(shape = 21, size = 4, alpha = 0.6)
```


## All about aesthetics, part 3

```{r, eval = F}
# Map cyl to size
ggplot(mtcars, aes(x = wt, y = mpg, size = cyl))+
geom_point()


# Map cyl to alpha
ggplot(mtcars, aes(x = wt, y = mpg, alpha = cyl)) +
geom_point()


# Map cyl to shape 
ggplot(mtcars, aes(x = wt, y = mpg, shape = cyl)) +
geom_point()


# Map cyl to label
ggplot(mtcars, aes(x = wt, y = mpg, label = cyl)) +
geom_point() +
geom_text()
```

## All about attributes, part 1

Notice that if an aesthetic and an attribute are set with the same argument, the attribute takes precedence. Once again, you see that the attribute needs to match the shape and geom, the fill aesthetic (or attribute) will only work with certain shapes.

```{r, eval = F}
# Define a hexadecimal color
my_color <- "#4ABEFF"

# Draw a scatter plot with color *aesthetic*
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl))+
geom_point()


# Same, but set color *attribute* in geom layer 
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
geom_point(col = my_color)

# Set the fill aesthetic; color, size and shape attributes
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl)) +
geom_point(size = 10, shape = 23, col = my_color)

```

## All about attributes, part 2

```{r, eval = F}
# Expand to draw points with alpha 0.5
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl)) +
  geom_point(alpha = 0.5)

  
# Expand to draw points with shape 24 and color yellow
gglpogt(mtcars, aes(x = wt, y = mpg, fill = cyl)) +
  geom_point(shape = 24, colour = "yellow")

# Expand to draw text with label rownames(mtcars) and color red
ggplot(mtcars, aes(x = wt, y = mpg, fill, label = rownames(mtcars), col = "red"))

```

## Going all out

```{r, eval = F}
# Map mpg onto x, qsec onto y and factor(cyl) onto col
ggplot(mtcars, aes(x = mpg, y = qsec, col = factor(cyl)))+
geom_point()


# Add mapping: factor(am) onto shape
ggplot(mtcars, aes(x = mpg, y = qsec, col = factor(cyl), shape = factor(am)))+
geom_point()


# Add mapping: (hp/wt) onto size
ggplot(mtcars, aes(x = mpg, y = qsec, col = factor(cyl), shape = factor(am), size = (hp/wt)))+
geom_point()
```

## Aesthetics for categorical and continuous variables

* Aesthetics that can be mapped onto continuous or categorical variable:
    * color & fill
    * alpha & size
    * alpha & label
    * x & y
    
* Some aesthetics are restricted to categorical data.
    
    * label & shape
    

```{r, eval = F}

```


## Position

```{r, eval = F}
# The base layer, cyl.am, is available for you
# Add geom (position = "stack" by default)
cyl.am + 
 geom_bar()

# Fill - show proportion
cyl.am + 
  geom_bar(position = "fill")  

# Dodging - principles of similarity and proximity
cyl.am +
  geom_bar(position = "dodge") 

# Clean up the axes with scale_ functions
val = c("#E41A1C", "#377EB8")
lab = c("Manual", "Automatic")
cyl.am +
  geom_bar(position = "dodge") +
  scale_x_discrete("Cylinders") + 
  scale_y_continuous("Number") +
  scale_fill_manual("Transmission", 
                    values = val,
                    labels = lab) 
```

## Setting a dummy aesthetic

```{r, eval = F}
# 1 - Create jittered plot of mtcars, mpg onto x, 0 onto y
ggplot(mtcars, aes(x = mpg, y = 0)) +
  geom_jitter()

# 2 - Add function to change y axis limits

ggplot(mtcars, aes(x = mpg, y = 0)) +
  scale_y_continuous(limits = c (-2, 2)) +
  geom_jitter()
```

## Aesthetics Best Pratices

* Aesthetics - Continuous Variables
     * x
     * y
     * size
     * alpha
     * colour
     * fill
     
* Aesthetics - Categorical Variables
     * labels
     * fill
     * shape
     * alpha
     * linetype
     * size
     
## Overplotting 1 - Point shape and transparency  
```{r, eval = F}
# Basic scatter plot: wt on x-axis and mpg on y-axis; map cyl to col
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl))+
geom_point(size = 4)


# Hollow circles - an improvement
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl))+
geom_point(size = 4, shape = 1)


# Add transparency - very nice
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl))+
geom_point(size = 4, alpha = 0.6)


```

## Overplotting 2 - alpha with large datasets

```{r, eval = F}
# Scatter plot: carat (x), price (y), clarity (color)
ggplot(diamonds, aes(x = carat, y = price, col = clarity))+
geom_point()

# Adjust for overplotting
ggplot(diamonds, aes(x = carat, y = price, col = clarity))+
geom_point(alpha = 0.5)


# Scatter plot: clarity (x), carat (y), price (color)
ggplot(diamonds, aes(x = clarity, y = carat, col = price)) +
geom_point(alpha = 0.5)


# Dot plot with jittering
ggplot(diamonds, aes(x = clarity, y = carat, col = price)) +
geom_point(alpha = 0.5, position = "jitter")
```

# 4. Geometries

## Scatter Plots and jittering (1)

jitter can be 1) an argument in geom_point(position = 'jitter'), 2) a geom itself, geom_jitter(), or 3) a position function, position_jitter(0.1)

```{r, eval = F}
# Shown in the viewer:
ggplot(mtcars, aes(x = cyl, y = wt)) +
  geom_point()

# Solutions:
# 1 - With geom_jitter()
ggplot(mtcars, aes(x = cyl, y = wt)) +
  geom_jitter()

# 2 - Set width in geom_jitter()
ggplot(mtcars, aes(x = cyl, y = wt)) +
  geom_jitter(width = 0.1)

# 3 - Set position = position_jitter() in geom_point() ()
ggplot(mtcars, aes(x = cyl, y = wt)) +
  geom_point(position = position_jitter(0.1))
```

## Scatter plots and jittering (2)

Notice how jittering and alpha blending serves as a great solution to the overplotting problem here. Setting the shape to 1 didn't really help, but it was useful in the previous exercises when you had less data. You need to consider each plot individually. You'll encounter this dataset again when you look at bar plots. In that case, treat each variable as categorical.

```{r, eval = F}
# Examine the structure of Vocab
str(Vocab)

# Basic scatter plot of vocabulary (y) against education (x). Use geom_point()
ggplot(Vocab, aes(x = education, y = vocabulary)) +
geom_point()


# Use geom_jitter() instead of geom_point()
ggplot(Vocab, aes(x = education, y = vocabulary)) +
geom_jitter(alpha = 0.2)


# Using the above plotting command, set alpha to a very low 0.2
ggplot(Vocab, aes(x = education, y = vocabulary)) +
geom_jitter(alpha = 0.2)


# Using the above plotting command, set the shape to 1
ggplot(Vocab, aes(x = education, y = vocabulary)) +
geom_jitter(shape = 1)
```

## Bar plots

* histogram, bar, errorbar

## Histograms

* x axis/aesthetics:
     * histograms cut up a continuous variable into discrete bins.

* y aestehtics:
     * only requires x aesthetics.
     * when geom_histogram()
     * ..density.. : proportional frequency of this bin in relation to the whole data set

```{r, eval = F}
# 1 - Make a univariate histogram

ggplot(mtcars, aes(x = mpg)) +
  geom_histogram()

# 2 - Plot 1, plus set binwidth to 1 in the geom layer
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 1)

# 3 - Plot 2, plus MAP ..density.. to the y aesthetic (i.e. in a second aes() function)
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(aes(y = ..density..), binwidth = 1)
```

## Position

* stack: place the bars on top of each other. Counts are used. This is the default position.
* fill: place the bars on top of each other, but this time use proportions.
* dodge: place the bars next to each other. Counts are used.

```{r, eval = F}
# Draw a bar plot of cyl, filled according to am
ggplot(mtcars, aes(x = cyl, fill = am)) +
  geom_bar()

# Change the position argument to stack
ggplot(mtcars, aes(x = cyl, fill = am)) +
  geom_bar(position = "stack")


# Change the position argument to fill
ggplot(mtcars, aes(x = cyl, fill = am)) +
  geom_bar(position = "fill")


# Change the position argument to dodge
ggplot(mtcars, aes(x = cyl, fill = am)) +
  geom_bar(position = "dodge")
```

## Overlapping bar plots

* the reason you want to use position_dodge() (and position_jitter()) is to specify how much dodging (or jittering) you want.


```{r, eval = F}
# 1 - The last plot form the previous exercise
ggplot(mtcars, aes(x = cyl, fill = am)) +
  geom_bar(position = "dodge")

# 2 - Define posn_d with position_dodge()
posn_d <- position_dodge(width = 0.2)

# 3 - Change the position argument to posn_d
ggplot(mtcars, aes(x = cyl, fill = am)) +
  geom_bar(position = posn_d)


# 4 - Use posn_d as position and adjust alpha to 0.6
ggplot(mtcars, aes(x = cyl, fill = am)) +
  geom_bar(position = posn_d, alpha = 0.6)

```

## Overlapping histograms

Outstanding overlapping! Frequency polygon plots, like kernel density plots, allow several distributions to be displayed in the same panel.

```{r, eval = F}
# A basic histogram, add coloring defined by cyl
ggplot(mtcars, aes(mpg, fill = cyl)) +
  geom_histogram(binwidth = 1)

# Change position to identity
ggplot(mtcars, aes(mpg, fill = cyl)) +
  geom_histogram(binwidth = 1, position = "identity")


# Change geom to freqpoly (position is identity by default)
ggplot(mtcars, aes(mpg, col = cyl)) +
  geom_freqpoly(binwidth = 1, position = "identity")
```

## Bar plots with color ramp, part 1

```{r, eval = F}
# Example of how to use a brewed color palette
ggplot(mtcars, aes(x = cyl, fill = am)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set1")

# Use str() on Vocab to check out the structure
str(Vocab)

# Plot education on x and vocabulary on fill
# Use the default brewed color palette
ggplot(Vocab, aes(x = education, fill = vocabulary)) +
geom_bar(position = "fill")+
scale_fill_brewer()
```

## Bar plots with color ramp, part 2

```{r, eval = F}
# Final plot of last exercise
ggplot(Vocab, aes(x = educatio, fill = vocabulary)) +
  geom_bar(position = "fill") +
  scale_fill_brewer()

# Definition of a set of blue colors
# from the RColorBrewer package
blues <- brewer.pal(9, "Blues")

# 1 - Make a color range using colorRampPalette() and the set of blues
blue_range <- colorRampPalette(blues)

# 2 - Use blue_range to adjust the color of the bars, use scale_fill_manual()

ggplot(Vocab, aes(x = education, fill = vocabulary)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = blue_range(11))

```


## Overlapping histograms (2)

```{r, eval = F}
# 1 - Basic histogram plot command
ggplot(mtcars, aes(mpg)) +
  geom_histogram(binwidth = 1)

# 2 - Plot 1, Expand aesthetics: am onto fill
ggplot(mtcars, aes(mpg, fill = am)) +
  geom_histogram(binwidth = 1)


# 3 - Plot 2, change position = "dodge"
ggplot(mtcars, aes(mpg, fill = am)) +
  geom_histogram(binwidth = 1, position = "dodge")



# 4 - Plot 3, change position = "fill"
ggplot(mtcars, aes(mpg, fill = am)) +
  geom_histogram(binwidth = 1, position = "fill")



# 5 - Plot 4, plus change position = "identity" and alpha = 0.4
ggplot(mtcars, aes(mpg, fill = am)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.4)


# 6 - Plot 5, plus change mapping: cyl onto fill
ggplot(mtcars, aes(mpg, fill = cyl)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.4)

```

## Line Plots 

```{r, eval = F}
# Print out head of economics
head(economics)

# Plot unemploy as a function of date using a line plot
ggplot(economics, aes(x = date, y = unemploy)) +
geom_line()


# Adjust plot to represent the fraction of total population that is unemployed
ggplot(economics, aes(x = date, y = unemploy/pop)) +
geom_line()
```


## Periods of recession

```{r, eval = F}

# Basic line plot
ggplot(economics, aes(x = date, y = unemploy/pop)) +
  geom_line()

# Expand the following command with geom_rect() to draw the recess periods
ggplot(economics, aes(x = date, y = unemploy/pop)) +
  geom_rect(data = recess,
         aes(xmin = begin, xmax = end, ymin = -Inf, ymax = +Inf), inherit.aes = FALSE, fill = "red", alpha = 0.2) +
  geom_line()
```

## Multiple time series, part 1

```{r, eval = F}
# Check the structure as a starting point
str(fish.species)

# Use gather to go from fish.species to fish.tidy
fish.tidy <- gather(fish.species, Species, Capture, -Year,)
```

## Multiple time series, part 2

```{r, eval = F}
# Recreate the plot shown on the right
ggplot(fish.tidy, aes(x = Year, y = Capture, col = Species)) +
geom_line()

```

# 5.qplot and wrap-up

## Using qplot

```{r, eval = F}
# The old way (shown)
plot(mpg ~ wt, data = mtcars) # formula notation
with(mtcars, plot(wt, mpg)) # x, y notation

# Using ggplot:
ggplot(mtcars, aes(wt, mpg)) +
  geom_point()

# Using qplot:
qplot(wt, mpg, data = mtcars)
```

## Using aesthetic

```{r, eval = F}
# basic qplot scatter plot:
qplot(wt, mpg, data = mtcars)

# Categorical variable mapped onto size:
# cyl
qplot(wt, mpg, data = mtcars, size = factor(cyl))

# gear
qplot(wt, mpg, data = mtcars, size = gear)

# Continuous variable mapped onto col:
# hp
qplot(wt, mpg, data = mtcars, col = hp)

# qsec
qplot(wt, mpg, data = mtcars, col = qsec)
```

## Choosing geoms, part 1

```{r, eval = F}
# qplot() with x only
qplot(data = mtcars, x = factor(cyl))

# qplot() with x and y
qplot(data = mtcars, x = factor(cyl), y = factor(vs))

# qplot() with geom set to jitter manually
qplot(data = mtcars, x = factor(cyl), y = factor(vs), geom = "jitter")
```

## Choosing geoms, part 2 - dotplot

* Scatter plots:
     * x : continuous;
     * y : continuous.

* Dot plots:
    * x : categorical;
    * y : continuous.

```{r, eval = F}
# cyl and am are factors, wt is numeric
class(mtcars$cyl)
class(mtcars$am)
class(mtcars$wt)

# "Basic" dot plot, with geom_point():
ggplot(mtcars, aes(cyl, wt, col = am)) +
  geom_point(position = position_jitter(0.2, 0)

# 1 - "True" dot plot, with geom_dotplot():               
ggplot(mtcars, aes(cyl, wt, fill = am)) +
  geom_dotplot(binaxis = "y", stackdir = "center")
 

# 2 - qplot with geom "dotplot", binaxis = "y" and stackdir = "center"
qplot(cyl, wt,
  data = mtcars,
  fill = am,
  geom = "dotplot",
  binaxis = "y",
  stackdir = "center"
)

# ChickWeight is available in your workspace
# 1 - Check out the head of ChickWeight
head(ChickWeight)

# 2 - Basic line plot
ggplot(ChickWeight, aes(x = Time, y = weight)) +
  geom_line(aes(group = Chick))

# 3 - Take plot 2, map Diet onto col.
ggplot(ChickWeight, aes(x = Time, y = weight, col = Diet)) +
  geom_line(aes(group = Chick))


# 4 - Take plot 3, add geom_smooth()
ggplot(ChickWeight, aes(x = Time, y = weight, col = Diet)) +
  geom_line(aes(group = Chick), alpha = 0.3)+
  geom_smooth(lwd = 2, se = F)
```

##Titanic

```{r, eval = F}
# titanic is avaliable in your workspace
# 1 - Check the structure of titanic
str(titanic)

# 2 - Use ggplot() for the first instruction
ggplot(titanic, aes(x = Pclass, fill = Sex)) +
  geom_bar(position = "dodge")

# 3 - Plot 2, add facet_grid() layer
ggplot(titanic, aes(x = Pclass, fill = Sex)) +
  geom_bar(position = "dodge") +
  facet_grid(. ~ Survived)


# 4 - Define an object for position jitterdodge, to use below
posn.jd <- position_jitterdodge(0.5, 0, 0.6)

# 5 - Plot 3, but use the position object from instruction 4
ggplot(titanic, aes(x = Pclass, y = Age, col = Sex)) +
  geom_point(size = 3, alpha = 0.5, position = posn.jd) +
  facet_grid(. ~ Survived)

```

