---
title: "Exploratory Data Analysis"
author: "Camila Bezerra"
date: "29/08/2019"
output:
  html_document:
    theme: united
    highlight: monochrome
---

```{r, echo=FALSE}
knitr::opts_chunk$set(eval = F)
```



# 1. Exploring Categorical Data

## Dropping levels

```{r, eval = F}
# Load dplyr
library(dplyr)

# Print tab  to find out which level of align has the fewest total entries.
tab

# Remove align level
comics_filtered <- comics %>%
  # Use filter() to filter out all rows of comics with that level
  filter(align != "Reformed Criminals") %>%
  #then drop the unused level with droplevels()
  droplevels()

# See the result
comics_filtered

```

## Side-by-side barcharts

You´ll construct two side-by-side barchats of the comics data. 
Passing the argument position = "dodge" to geom_bar() says that yu want a side-by-side (i.e. not stacked) barchart.

```{r, eval = F}
# Load ggplot2
library(ggplot2)

# Create side-by-side barchart of gender by alignment
ggplot(comics, aes(x = align, fill = gender)) + 
  geom_bar(position = "dodge")

# Create side-by-side barchart of alignment by gender
ggplot(comics, aes(x = gender, fill = align)) + 
  geom_bar(position = "dodge") +
  #Rotate the axis labels 90 degrees to help readability.
  theme(axis.text.x = element_text(angle = 90))
```


## Conditional proportions

```{r, eval = F}
tab <- table(comics$align, comics$gender)
options(scipen = 999, digits = 3) # Print fewer digits
prop.table(tab)     # Joint proportions
prop.table(tab, 2)  # Conditional on columns
```



## Count vs. proportions(2)

Bar charts can tell dramatically different stories depending on whether they represent counts or proportions and, if proportions, what the proportions are conditioned on. To demonstrate this difference, you'll construct two barcharts in this exercise: one of counts and one of proportions.

```{r, eval = F}
# Plot of gender by align
ggplot(comics, aes(x = align, fill = gender)) +
  geom_bar()
  
# Plot proportion of gender, conditional on align
ggplot(comics, aes(x = align, fill = gender)) + 
geom_bar(position = "fill") +
  ylab("proportion")
```


## Marginal barchart

```{r, eval = F}
# Change the order of the levels in align
comics$align <- factor(comics$align, 
                       levels = c("Bad", "Neutral", "Good"))

# Create plot of align
ggplot(comics, aes(x = align)) + 
  geom_bar()
```


## Conditional barchart

If you want to break down the distribution of alignment based on gender, you're looking for conditional distributions.

```{r, eval = F}
# Plot of alignment broken down by gender
ggplot(comics, aes(x = align)) + 
  geom_bar() +
  facet_wrap(~ gender)
```


# 2. Exploring Numerical Data

## Boxplots and density plots

```{r, eval = F}
# Filter cars with 4, 6, 8 cylinders
common_cyl <- filter(cars,ncyl %in% c(4, 6, 8))

# Create box plots of city mpg by ncyl
ggplot(common_cyl, aes(x = as.factor(ncyl), y = city_mpg)) +
  geom_boxplot()

# Create overlaid density plots for same data
ggplot(common_cyl, aes(x = city_mpg, fill = as.factor(ncyl))) +
  geom_density(alpha = .3)
```


## Choose measures for center and spread

Like mean and standard deviation, median and IQR measure the central tendency and spread, respectively, but are robust to outliers and non-normal data.

## Transformations

```{r, eval = F}
# Create density plot of old variable
gap2007 %>%
  ggplot(aes(x = pop)) +
  geom_density()

# Transform the skewed pop variable
gap2007 <- gap2007 %>%
  mutate(log_pop = log(pop))

# Create density plot of new variable
gap2007 %>%
  ggplot(aes(x = log_pop)) +
  geom_density()
```

## Outliers

### Characteristics of a distribution

* Center
* Variability
* Shape 
* Outliers

### Identify outliers

```{r, eval = F}
# Filter for Asia, add column indicating outliers
gap_asia <- gap2007 %>%
  filter(continent == "Asia") %>%
  mutate(is_outlier = lifeExp < 50)

# Remove outliers, create box plot of lifeExp
gap_asia %>%
  filter(!is_outlier) %>%
  ggplot(aes(x = 1, y = lifeExp)) +
  geom_boxplot()
```
