---
title: "Exploratory Data Analysis - Case Study"
author: "Camila Bezerra"
date: "30/08/2019"
output:
  html_document:
    theme: united
    highlight: monochrome
---

```{r, echo=FALSE}
knitr::opts_chunk$set(eval = F)
```



# 1. Data cleaning and summarizing with dplyr

## Adding a country column

```{r, eval = F}
# Load the countrycode package
library(countrycode)

# Convert country code 100
countrycode(100, "cown", "country.name")

# Add a country column within the mutate: votes_processed
votes_processed <- votes %>%
  filter(vote <= 3) %>%
  mutate(year = session + 1945,
 country = countrycode(ccode, "cown", "country.name"))
```


## Summarizing the full dataset

summarize(): turns many rows into one

```{r, eval = F}
# Print votes_processed
votes_processed

# Find total and fraction of "yes" votes
votes_processed %>% 
  summarize(total = n(),
            percent_yes = mean(vote == 1))
```

# 2. Data visualization with ggplot2


## Plotting a line over time

```{r, eval = F}
# Define by_year
by_year <- votes_processed %>%
  group_by(year) %>%
  summarize(total = n(),
            percent_yes = mean(vote == 1))

# Load the ggplot2 package
library(ggplot2)

# Create line plot
ggplot(by_year, aes(x = year, y = percent_yes)) +
  geom_line()
```


## Other ggplot2 layers


```{r, eval = F}
# Change to scatter plot and add smoothing curve
ggplot(by_year, aes(year, percent_yes)) +
  geom_point() +
  geom_smooth()
```


## Plotting multiple countries

```{r, eval = F}
# Vector of four countries to examine
countries <- c("United States", "United Kingdom",
               "France", "India")

# Filter by_year_country: filtered_4_countries
filtered_4_countries <- by_year_country %>%
filter(country %in% countries)

# Line plot of % yes in four countries
ggplot(filtered_4_countries, aes(x = year, y = percent_yes, color = country)) +
  geom_line()
```


## Faceting with free y-axis

```{r, eval = F}
# Vector of six countries to examine
countries <- c("United States", "United Kingdom",
               "France", "Japan", "Brazil", "India")

# Filtered by_year_country: filtered_6_countries
filtered_6_countries <- by_year_country %>%
  filter(country %in% countries)

# Line plot of % yes over time faceted by country
ggplot(filtered_6_countries, aes(year, percent_yes)) +
  geom_line() +
  facet_wrap(~ country, scales = "free_y")
```

# 3. Tidy modeling with broom

## Linear regression

```{r, eval = F}
# Percentage of yes votes from the US by year: US_by_year
US_by_year <- by_year_country %>%
  filter(country == "United States")

# Print the US_by_year data
US_by_year

# Perform a linear regression of percent_yes by year: US_fit
US_fit <- lm(formula = percent_yes ~ year, data = US_by_year)

# Perform summary() on the US_fit object
summary(US_fit)
```

# 3. Tidy modeling with broom

## Tidying models with broom
 
### broom turns a model into a data frame

```{r, eval = F}
library(broom)
tidy(model)
```

## Combining models for mutiple countries

```{r, eval = F}
# Linear regression of percent_yes by year for US
US_by_year <- by_year_country %>%
  filter(country == "United States")
US_fit <- lm(percent_yes ~ year, US_by_year)

# Fit model for the United Kingdom
UK_by_year <- by_year_country %>%
filter(country == "United Kingdom")
UK_fit <- lm(percent_yes ~ year, UK_by_year)

# Create US_tidied and UK_tidied
US_tidied <- tidy(US_fit)
UK_tidied <- tidy(UK_fit)


# Combine the two tidied models
bind_rows(US_tidied, UK_tidied)
```


## Nesting for multiple models

* nest() turns it into one row per country

```{r, eval = F}
library(tidyr)
by_year_country %>% 
  nest(-country) # means "nest all except country"

```

* unnest() does the opposite

```{r, eval = F}
by_year_country %>% 
  nest(-country) %>% # means "nest all except country"
  unnest(data)
```

## List comlumns

This "nested" data has an interesting structure. The second column, data, is a list, a type of R object that hasn't yet come up in this course that allows complicated objects to be stored within each row. This is because each item of the data column is itself a data frame.

```{r, eval = F}
# All countries are nested besides country
nested <- by_year_country %>%
  nest(-country)

# Print the nested data for Brazil
nested$data[[7]]
```


## Fitting multiple countries

* map() applies an operation to each item in a list

```{r}
v <- list(1, 2, 3)
map(v, ~. * 10)
```


```{r, eval = F}
library(purrr)
by_year_country %>% 
  nest(-country) %>% 
#map() fits a model to each dataset
  mutate(models = map(data, ~lm(percent_yes ~year, data = .))) %>% #tidy turns each model into a data frame
  mutate(tidied ~ map(models, tidy)) %>% 
#unnest(combines the tidied models)
unnest(tidied)
```


## Performing linear regression on each nested dataset

```{r, eval = F}
# Perform a linear regression on each item in the data column
by_year_country %>% 
  group_by(country) %>% # agrupar por país
  nest() %>%  
  mutate(model = map(data, ~lm(percent_yes ~year, data = . ))) #aplicar o modelo para cada país
```

## Tidy each linear regression model

```{r, eval = F}
# Load the broom package
library(broom)

# Add another mutate that applies tidy() to each model
by_year_country %>%
  nest(-country) %>%
  mutate(
  model = map(data, ~ lm(percent_yes ~ year, data = .)),
  tidied = map(model, tidy)
  )
```


## Working with many tidy models

```{r, eval = F}
country_coefficients %>% 
  filter(term == "year") %>% # filter for the year term(slope)
  filter(p.adjust(p.value) < .05) #filtered by adjusted p-value
```



## Filtering for significant countries

Use the p.adjust() function to adjust the p.value column, saving the result into a new p.adjusted column. Then, filter for cases where p.adjusted is less than .05.

```{r, eval = F}
# Filter for only the slope terms
slope_terms <- cuntry_coefficients %>% 
  filter(term == "year")

# Add p.adjusted column, then filter 
slope_terms %>% 
  mutate(p.adjusted = p.adjust(p.value)) %>% 
  filter(p.adjusted < .05)
```

# 4. Joining and tidying

## Tidy data

* Use gather() to bring columns into two variables

```{r, eval = F}
library(tidyr)
votes_joined %>% 
  gather(topic, has_topic, me:ec)

# "topic" is now a variable
```


## Using gather to tidy dataset

In order to represent the joined vote-topic data in a tidy form so we can analyze and graph by topic, we need to transform the data so that each row has one combination of country-vote-topic. This will change the data from having six columns (me, nu, di, hr, co, ec) to having two columns (topic and has_topic).

```{r, eval = F}
# Load the tidyr package
library(tidyr)

# Gather the six me/nu/di/hr/co/ec columns
votes_joined %>% gather(topic, has_topic, me:ec)


# Perform gather again, then filter
votes_gathered <- votes_joined %>% gather(topic, has_topic, me:ec)%>%
filter(has_topic == 1)

```

## Summarize by country, year, and topic

```{r, eval = F}
# Print votes_tidied
votes_tidied

# Summarize the percentage "yes" per country-year-topic
by_country_year_topic <- votes_tidied %>%
group_by(country, year, topic) %>%
summarize(total = n(),
percent_yes = mean(vote == 1)) %>%
ungroup()

# Print by_country_year_topic
by_country_year_topic
```


## Tidy modeling by topic and country

```{r, eval = F}

library(purr)
library(broom)

country_topic_coefficients <- by_year_country_topic %>% 
  nest(~country, -topic) %>% 
  mutate(model = map(data, ~lm(percent_yes ~ year, data = .)),
         tidied = map(model, tidy)) %>% 
  unnest(tidied)

```

##

```{r, eval = F}

```