---
title: "Introduction to the Tidyverse"
author: "Camila Bezerra"
date: "01/08/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo = F, warning=FALSE}
library(tidyverse)
library(dplyr)
library(utils)
library(readr)
```

# I- Importing data in R

## Introduction & read.csv

1. Flat Files
*csv: Comma Separated Values

```{r, eval=FALSE}

#se o arquivo estiver no seu working directory
read.csv("states.csv")

#What if file in datasets folder of home directory?
path <- file.path("~", "datasets", "states.csv")

path

dir() #para ver as pastas presentes no working directory
```


```{r, eval=FALSE}
# Import swimming_pools.csv: pools
pools <- read.csv("swimming_pools.csv")

# Print the structure of pools
str(pools)
```

stringsAsFactors: tells R whether it should convert strings in the flat file to factors.
stringsAsFactors = T, means that you import strings as factors
stringsAsFactors = F, the data frame columnns corresponding to strings in your text fiel will be character.

```{r, eval =FALSE}
# Import swimming_pools.csv correctly: pools
pools <- read.csv("swimming_pools.csv", stringsAsFactors = F)
#stringAsFactors = F. POrque nao queremos que as variaveis do data.frame sejam convertidas em fatores.

# Check the structure of pools
str(pools)
```

2. read.delim & read.table

*.txt : Tab-delimed file

read.table()
*read any tabular file as a data frame

```{r, eval=FALSE}

read.table("states.txt",
           header = T, #first row lists variable names(default False)
           sep = "/", #field separator is a forward slash
           stringsAsFactors = F)
```

```{r, eval = FALSE}
# Import hotdogs.txt: hotdogs
hotdogs <- read.table("hotdogs.txt",
header = F)

# Summarize hotdogs
summary(hotdogs)
```

```{r, eval = F}
# Path to the hotdogs.txt file: path
path <- file.path("data", "hotdogs.txt") #salvar path em um objeto

# Import the hotdogs.txt file: hotdogs
hotdogs <- read.table(path, #objeto salvo
                      sep = "\t", #separar por tab
                      col.names = c("type", "calories", "sodium")) #definir os nomes das colunas, uma vez que no arquivo a primeira linha nao contem os nomes.
#header = F, por default

# Call head() on hotdogs
head(hotdogs)
```

```{r, eval = F}
# Finish the read.delim() call
hotdogs <- read.delim("hotdogs.txt", header = F, col.names = c("type", "calories", "sodium"))

# Select the hot dog with the least calories: lily
lily <- hotdogs[which.min(hotdogs$calories), ]

# Select the observation with the most sodium: tom
tom <- hotdogs[which.max(hotdogs$sodium), ]

# Print lily and tom
lily
tom
```


colClasses
*specify the columnn types or columnn classes of the resulting data frame.

```{r, eval = F}
read.delim("my_file.txt",
           colClasses = c("character",
                          "numeric",
                          "logical"))
```

*If a columnn is set to "NULL" in the colClasses vector, this columnn will be skipped and will not be loaded into the data frame.

## Wrappers

*read.table(): the main function
*read.csv() = wrapper for CSV
*read.delim() = wrapper for tab-delimited files

## read.csv

* Defaults:
** header = T
** sep = ","

##read.csv2
** sep = ";"

##read.delim

* Defaults
** header = T
** sep = "\t"

##read.delim2 

* sep = "\t"

# II - readr & data.table

## readr package

```{r, eval = F}
read_csv("states.csv")

#when it prints the results, it shows the columns classes


read_tsv("states.txt")
```


## readr: read_delim

```{r, eval = F}
read_delim("states2.txt", delim = "/")
```

* col_types
** to control the columns classes


```{r, eval = F}
read_delim("states2.txt", delim = "/t", col_types = "ccddd")
#c = character
#d = double
#i = integer
#l = logical
#- = skip
```

* skip and n_max
** to skip rows and read a limited number of rows
** skip: specifies the number of lines you're ignoring in the flat file before actually starting to import data.
** n_max: specifies the number of lines you're actually importing.


```{r, eval = F}
read_delim("states2.txt", delim = "/t",
           col_names = c("State", "city", "pop", "area"),
           skip = 2, n_max = 3) #pular 2 linhas, ler as proximas 3.
```

## col_types with collectors

* Collector functions can be passes in a list() to the col_types argument of read_ functions to tell them how to interpret values in a column.

** col_integer(): the column should be interpreted as an integer
** col_factor(levels, orderes = F):
the column should be interpreted as a fatcor with levels.

```{r, eval = F}

# readr is already loaded

# Import without col_types
hotdogs <- read_tsv("hotdogs.txt", col_names = c("type", "calories", "sodium"))

# Display the summary of hotdogs
summary(hotdogs)

# The collectors you will need to import the data
fac <- col_factor(levels = c("Beef", "Meat", "Poultry"))
int <- col_integer()

# Edit the col_types argument to import the data correctly: hotdogs_factor
hotdogs_factor <- read_tsv("hotdogs.txt",col_names = c("type", "calories", "sodium"), col_types = list(fac, int, int))

# Display the summary of hotdogs_factor
summary(hotdogs_factor)
```


## data.table:fread

* data table package
* fread() function:
** similar to read.table() functions
** infer colnames, column types and separators

```{r, eval = F}
# fread is already loaded

# Import columns 6 and 8 of potatoes.csv: potatoes
potatoes <- fread("potatoes.csv", select = c("texture", "moistness"))

# Plot texture (x) and moistness (y) of potatoes
plot(potatoes$texture,potatoes$moistness)
```

 # 3. readxl(1)
 
 * excel_sheets(): list different sheets
 
```{r,eval=F}
dir() 
# [1] "cities.xlsx" "the_rest_is_secret.txt"

excel_sheets("cities.xlsx")
# [1] "year_1990" "year_2000"
```
 
 * read_excel(): actually import data into R
 
```{r,eval=F}
read_excel("cities.xlsx")
#by defualt the first sheet is imported

#para especificar a pasta
read_excel("cities.xlsx", sheet = 2)

#ou

read_excel("cities.xlsx", sheet = "year_2000")

```
 
```{r, eval = F}
# The readxl package is already loaded

# Read the sheets, one by one
pop_1 <- read_excel("urbanpop.xlsx", sheet = 1)
pop_2 <- read_excel("urbanpop.xlsx", sheet = 2)
pop_3 <- read_excel("urbanpop.xlsx", sheet = 3)

# Put pop_1, pop_2 and pop_3 in a list: pop_list
pop_list <- list(pop_1, pop_2, pop_3)

# Display the structure of pop_list
str(pop_list)

```

Aplicando a funcao lapply para ler varias sheets de uma vez

```{r, eval = F}
# The readxl package is already loaded

# Read all Excel sheets with lapply(): pop_list
#aplicar a funcao lapply
#especificar excel_sheets para escolher as sheets
#ler as sheets
pop_list <- lapply(excel_sheets("urbanpop.xlsx"),
read_excel,
path="urbanpop.xlsx")

# Display the structure of pop_list
str(pop_list)
```


## readxl(2)

*col_names

```{r, eval = F}
# The readxl package is already loaded

# Import the first Excel sheet of urbanpop_nonames.xlsx (R gives names): pop_a
pop_a <- read_excel("urbanpop_nonames.xlsx", col_names= F)

# Import the first Excel sheet of urbanpop_nonames.xlsx (specify col_names): pop_b
cols <- c("country", paste0("year_", 1960:1966))

pop_b <- read_excel("urbanpop_nonames.xlsx",
col_names = cols)

# Print the summary of pop_a
summary(pop_a)

# Print the summary of pop_b
summary(pop_b)
```

*skip argument

```{r, eval = F}
# The readxl package is already loaded

# Import the second sheet of urbanpop.xlsx, skipping the first 21 rows: urbanpop_sel
urbanpop_sel <- read_excel("urbanpop.xlsx", sheet=2, col_names = F, skip = 21)

# Print out the first observation from urbanpop_sel
urbanpop_sel[1, ]

```

## gdata package

*read.xls()

```{r, eval = F}
# Load the gdata package
library(gdata)

# Import the second sheet of urbanpop.xls: urban_pop
urban_pop <- read.xls("urbanpop.xls", sheet = "1967-1974")

# Print the first 11 observations using head()
head(urban_pop, n = 11)

# The gdata package is alreaded loaded

# Column names for urban_pop
columns <- c("country", paste0("year_", 1967:1974))

# Finish the read.xls call
urban_pop <- read.xls("urbanpop.xls", sheet = 2,
                      skip = 50, header = F, stringsAsFactors = F,
                      col.names = columns)

# Print first 10 observation of urban_pop
head(urban_pop, n = 10)

# Add code to import data from all three sheets in urbanpop.xls
path <- "urbanpop.xls"
urban_sheet1 <- read.xls(path, sheet = 1, stringsAsFactors = FALSE)
urban_sheet2 <- read.xls(path, sheet = 2, stringsAsFactors = FALSE)
urban_sheet3 <- read.xls(path, sheet = 3, stringsAsFactors = F)


# Extend the cbind() call to include urban_sheet3: urban
urban <- cbind(urban_sheet1, urban_sheet2[-1], urban_sheet3[-1])

# Remove all rows with NAs from urban: urban_clean
urban_clean <- na.omit(urban)


# Print out a summary of urban_clean
summary(urban_clean)
```

## Reproducible Excel work with XLConnect

#XLConnect

```{r, eval=F}
library("XLConnect")

book <- loadWorkbook("cities.xlsx")

str(book)

getSheets(book)#para pegar as sheets

#para importar uma sheet

readWorksheet(book, sheet = "year_2000")


# XLConnect is already available

# Build connection to urbanpop.xlsx
my_book <- loadWorkbook("urbanpop.xlsx")

# Import columns 3, 4, and 5 from second sheet in my_book: urbanpop_sel
urbanpop_sel <- readWorksheet(my_book, sheet = 2, startCol = 3, endCol = 5)

# Import first column from second sheet in my_book: countries
countries <- readWorksheet(my_book, sheet = 2, startCol = 1, endCol = 1)

# cbind() urbanpop_sel and countries together: selection
selection <- cbind(countries, urbanpop_sel )

# XLConnect is already available

# Build connection to urbanpop.xlsx
my_book <- loadWorkbook("urbanpop.xlsx")

# Add a worksheet to my_book, named "data_summary"
createSheet(my_book, "data_summary")

# Create data frame: summ
sheets <- getSheets(my_book)[1:3]
dims <- sapply(sheets, function(x) dim(readWorksheet(my_book, sheet = x)), USE.NAMES = FALSE)
summ <- data.frame(sheets = sheets,
                   nrows = dims[1, ],
                   ncols = dims[2, ])

# Add data in summ to "data_summary" sheet
writeWorksheet(my_book, summ, "data_summary")

# Save workbook as summary.xlsx
saveWorkbook(my_book, "summary.xlsx")
```

*renameSheet

```{r, eval = F}
# my_book is available

# Rename "data_summary" sheet to "summary"
renameSheet(my_book, "data_summary", "summary")

# Print out sheets of my_book
getSheets(my_book)

# Save workbook to "renamed.xlsx"
saveWorkbook(my_book,"renamed.xlsx")
```

