---
title: "Unsupervised Learning in R"
author: "Camila Bezerra"
date: "10/09/2019"
output:
  html_document:
    theme: united
    highlight: monochrome
---

```{r, warning=F}
library(ggplot2)
library(tidyverse)
library(dplyr)
```

# 1. Introduction to Factor Variables

## Summarising factors

```{r}
#get the number of categories(levels)
nlevels(xxx)

#get the list of categories (levels)
nlevels()

# get number of levels for every factor variable
multipleChoiceResponses %>% 
  summarise_if(is.factor, nlevels)
```

## Getting number of levels

There are more handy functions - summarise_at() and mutate_at(). These let you select columns by name-based select helpers.

```{r}
# Change all the character columns to factors
responses_as_factors <- multiple_choice_responses %>%
    mutate_if(is.character, as.factor)

number_of_levels <- responses_as_factors %>%
	# apply the function nlevels to each column
    summarise_all(nlevels) %>%
    # change the dataset from wide to long
    gather(variable, num_levels)
```

## Examining number of levels

```{r}
# Select the 3 rows with the highest number of levels
number_of_levels %>%
    top_n(3, num_levels)
    
number_of_levels %>%
	# filter for where variable equals CurrentJobTitleSelect
    filter(variable == "CurrentJobTitleSelect") %>%
	# pull num_levels
    pull()
```

## Examining levels

```{r}
responses_as_factors %>%
    # pull CurrentJobTitleSelect
    pull(CurrentJobTitleSelect) %>%
    # get the values of the levels
    levels()
```

## Making better plots

```{r}
## Reordering factors
ggplot(workChallenges) +
  geom_point(aes(x = fct_reorder(question, perc_problem), y = perc_problem))

## reordering bar chart
ggplot(multiple_choice_responses)+
  geom_bar(aes(x = fct_infreq(CurrentJobTitleSelect)))

## reversing factor levels
ggplot(multiple_choice_responses)+
  geom_bar(aes(x = fct_rev(fct_infreq(CurrentJobTitleSelect))))
```

## Reordering a variable by its frequency

```{r}
# Make a bar plot
ggplot(multiple_choice_responses, aes(x =EmployerIndustry)) + 
    geom_bar() + 
    # flip the coordinates
    coord_flip()

# Make a bar plot
ggplot(multiple_choice_responses, aes(x =fct_rev(fct_infreq(EmployerIndustry)))) + geom_bar() + 
    # flip the coordinates
    coord_flip()
```

## Ordering one variable by another

Congrats, you've finished the chapter! You've now learned how to identify qualitative variables, get the number and names of the levels, and reorder variables.

```{r}
multiple_choice_responses %>%
  # remove NAs
  filter(!is.na(EmployerIndustry) & !is.na(Age)) %>%
  # get mean_age by EmployerIndustry
  group_by(EmployerIndustry) %>%
  summarise(mean_age = mean(Age)) %>%
  # reorder EmployerIndustry by mean_age 
  mutate(EmployerIndustry = fct_reorder(EmployerIndustry, mean_age)) %>%
  # make a scatterplot of EmployerIndustry by mean_age
  ggplot(aes(x = EmployerIndustry, y = mean_age)) + 
    geom_point() + 
    coord_flip()
```

# 2. Manipulating Factor Variables

## Reordering factors

```{r}
ggplot(aes(nlp_frquency,
           x = fct_relevel(response,
                         "Rarely", "Sometimes", "Often", "Most of the time"))) +
  geom_bar()
```

### fct_reorder()

```{r}
nlp_frequency %>% 
  pull(response) %>% 
  levels()

nlp_frequency %>% 
  mutate(response = fct_relevel(response,
                                "Often", "Most of the time")) %>% 
  pull(response) %>% 
  levels()
```

### Additional arguments

```{r}

nlp_frequency %>% 
  mutate(response = fct_relevel(response,
                                "Often", "Most of the time", after = 2)) %>% 
  pull(response) %>% 
  levels()

nlp_frequency %>% 
  mutate(response = fct_relevel(response,
                                "Often", "Most of the time", after = Inf)) %>% 
  pull(response) %>% 
  levels()
```


## Changing the order of factor levels

```{r}
# Get the levels of WorkInternalVsExternalTools
print(multiple_choice_responses$WorkInternalVsExternalTools)

# Reorder the levels from internal to external 
mc_responses_reordered <- multiple_choice_responses %>%
    mutate(WorkInternalVsExternalTools = fct_relevel(WorkInternalVsExternalTools,
                                            "Entirely internal", 
                                             "More internal than external",
                                             "Approximately half internal and half external",
                                             "More external than internal", 
                                             "Entirely external",
                                             "Do not know"))

# Make a bar plot of the responses
ggplot(mc_responses_reordered, aes(x = WorkInternalVsExternalTools)) + 
    geom_bar() + 
    coord_flip()
```

## Tricks of fct_relevel()

```{r}
multiple_choice_responses %>%
    # Move "I did not complete any formal education past high school" and "Some college/university study without earning a bachelor's degree" to the front
    mutate(FormalEducation = fct_relevel(FormalEducation, "I did not complete any formal education past high school", "Some college/university study without earning a bachelor's degree")) %>%
    # Move "I prefer not to answer" to be the last level.
    mutate(FormalEducation = fct_relevel(FormalEducation, "I prefer not to answer", after = Inf)) %>%
    # Move "Doctoral degree" to be after the 5th level
    mutate(FormalEducation = fct_relevel(FormalEducation, "Doctoral degree", after = 5)) %>%
    # Examine the new level order
    pull(FormalEducation) %>%
    levels()
```

## Renaming unknown levels

```{r}
flying_etiquette %>% 
    mutate(middle_arm_rest_three = fct_recode(middle_arm_rest_three,
    "Everyone should share" = "arm rests should be share")) %>%
     count(middle_arm_rest_three)
```

fct_recode() is for changing the names of the levels.

## Renaming a few levels

```{r}
# make a bar plot of the frequency of FormalEducation
ggplot(multiple_choice_responses, aes(x = FormalEducation)) + 
    geom_bar()

multiple_choice_responses %>%
    # rename levels
    mutate(FormalEducation = fct_recode(FormalEducation, 
    "High school" = "I did not complete any formal education past high school", 
    "Some college" = "Some college/university study without earning a bachelor's degree")) %>%
    # make a bar plot of FormalEducation
    ggplot(aes(x = FormalEducation)) + 
    geom_bar()
```

## Collapsing factor levels

fct_collpase()

```{r}
flying_etiquette %>%
  mutate(height = fct_collapse(height, 
             under_5_3 = c("Under 5 ft.", "5'0\"", "5'1\"", "5'2\""),
             over_6_1 = c("6'1\"", "6'2\"", "6'3\"", "6'4\"", 
             "6'5\"", "6'6\" and above"))) %>%
  pull(height) %>%
  levels()
```

fct_other() keep

```{r}
flying_etiquette %>%
  mutate(new_height = fct_other(height, keep = c("6'4\"", "5'1\""))) %>%
  count(new_height)
```

fct_other() drop

```{r}
flying_etiquette %>%
  mutate(new_height = fct_other(height, 
  drop = c("Under 5 ft.", "5'0\"", "5'1\"", "5'2\"", "5'3\""))) %>%
  pull(new_height) %>%
  levels()
```

fct_lump() prop

```{r}
flying_etiquette %>%
  mutate(new_height = fct_lump(height, prop = .08)) %>%
  count(new_height)
```

## Manually collapsing levels

```{r}
multiple_choice_responses %>%
    # Create new variable, grouped_titles, by collapsing levels in CurrentJobTitleSelect
    mutate(grouped_titles = fct_collapse(CurrentJobTitleSelect, 
        "Computer Scientist" = c("Programmer", "Software Developer/Software Engineer"), 
        "Researcher" = "Scientist/Researcher", 
        "Data Analyst/Scientist/Engineer" = c("DBA/Database Engineer", "Data Scientist", 
                                              "Business Analyst", "Data Analyst", 
                                              "Data Miner", "Predictive Modeler"))) %>%
    # Turn every title that isn't now one of the grouped_titles into "Other"
    mutate(grouped_titles = fct_other(grouped_titles, 
                             keep = c("Computer Scientist", 
                                     "Researcher", 
                                     "Data Analyst/Scientist/Engineer"))) %>% 
    # Get a count of the grouped titles
    count(grouped_titles)
```

## Lumping variables by proportion

```{r}
multiple_choice_responses %>%
  # remove NAs of MLMethodNextYearSelect
  filter(!is.na(MLMethodNextYearSelect)) %>%
  # create ml_method, which lumps all those with less than 5% of people into "Other"
  mutate(ml_method = fct_lump(MLMethodNextYearSelect, prop = .05)) %>%
  # count the frequency of your new variable, sorted in descending order
  count(ml_method, sort = TRUE)
```

## Preserving the most common levels

```{r}
multiple_choice_responses %>%
  # remove NAs 
  filter(!is.na(MLMethodNextYearSelect)) %>%
  # create ml_method, retaining the 5 most common methods and renaming others "other method" 
  mutate(ml_method = fct_lump(MLMethodNextYearSelect, n = 5, other_level = "other method")) %>%
  # count the frequency of your new variable, sorted in descending order
  count(ml_method, sort = TRUE)
```


# 3. Creating Factor Variables

## Examining common themed variables

```{r}
# Selecting and gathering data

multipleChoiceResponses %>% 
  select(contains("WorkChallendeFrequency")) %>% 
  gather(work_challenge, frequency)

## Changing strings
work_challenges <- multipleChoiceResponses %>% 
  select(contains("WorkChallengeFrequency")) %>% 
  gather(work_challenge, frequency) %>% 
  mutate(work_challenge = str_remove(work_challenge,
                                     "WorkChallengeFrequency"))

## if_else() and summarizing
work_challenges %>% 
  filter(!is.na(frequency)) %>% 
  mutate(frequency = if_else(
    frequency %in% c ("Most of the time", "Often"),
    1,
    0)
  ) %>% 
  group_by(work_challenge) %>% 
  summarise(perc_problem = mean(frquency))
```

## Grouping and reshaping similar columns

```{r}
learning_platform_usefulness <- multiple_choice_responses %>%
  # select columns with LearningPlatformUsefulness in title
  select(contains("LearningPlatformUsefulness")) %>%
  # change data from wide to long
  gather(learning_platform, usefulness) %>%
  # remove rows where usefulness is NA
  filter(!is.na(usefulness)) %>%
  # remove "LearningPlatformUsefulness" from each string in learning_platform 
  mutate(learning_platform = str_remove(learning_platform, "LearningPlatformUsefulness"))
```

## Summarizing data

```{r}

learning_platform_usefulness %>%
  # change dataset to one row per learning_platform usefulness pair with number of entries for each
  count(learning_platform, usefulness)

learning_platform_usefulness %>%
  # change dataset to one row per learning_platform usefulness pair with number of entries for each
  count(learning_platform, usefulness) %>%
  # use add_count to create column with total number of answers for that learning_platform
  add_count(learning_platform, wt = n)


perc_useful_platform <- learning_platform_usefulness %>%
  # change dataset to one row per learning_platform usefulness pair with number of entries for each
  count(learning_platform, usefulness) %>%
  # use add_count to create column with total number of answers for that learning_platform
  add_count(learning_platform, wt = n) %>%
  mutate(perc = learning_platform)

perc_useful_platform <- learning_platform_usefulness %>%
  # change dataset to one row per learning_platform usefulness pair with number of entries for each
  count(learning_platform, usefulness) %>%
  # use add_count to create column with total number of answers for that learning_platform
  add_count(learning_platform, wt = n) %>%
  # create a new column, perc, that is the percentage of people giving that response for that learning_platform
  mutate(perc = n / nn)

# create a line graph for each question with usefulness on x-axis and percentage of responses on y
ggplot(perc_useful_platform, aes(x = usefulness, y = perc, group = learning_platform)) + 
  geom_line() + 
  facet_wrap(~ learning_platform)
```

## Creating an initial plot

```{r}
# Make a scatter plot of average usefulness by learning platform 
ggplot(usefulness_by_platform, aes(x = learning_platform, y = avg_usefulness)) + 
        geom_point()
```

## Tricks of ggplot2

```{r}
## changing tick labels angle

ggplot(job_titles_by_perc,
       aes(x = CurrentJobTitleSelect, y = perc_w_title))+
  geom_point()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

## Using fct_reorder()

ggplot(job_titles_perc, aes(x = fct_reorder(CurrentJobTitleSelect,
                                            perc_w_title), y = perc_w_title))+
  geom_point()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

## Using labs()
ggplot(job_titles_by_perc, 
        aes(x = fct_rev(fct_reorder(CurrentJobTitleSelect, perc_w_title)),
            y = perc_w_title)) + 
    geom_point() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(x = "Job Title", y = "Percent with title")

## Changing to % scales
ggplot(job_titles_by_perc, 
        aes(x = fct_rev(fct_reorder(CurrentJobTitleSelect, perc_w_title)),
        y = perc_w_title)) + 
    geom_point() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(x = "Job Title", y = "Percent with title") + 
    scale_y_continuous(labels = scales::percent_format())
```

## Editing plot text

```{r}
ggplot(usefulness_by_platform, aes(x = learning_platform, y = avg_usefulness)) + 
    geom_point() + 
    # rotate x-axis text by 90 degrees
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    # rename y and x axis labels
    labs(x = "Learning Platform", y = "Percent finding at least somewhat useful") + 
    # change y axis scale to percentage
    scale_y_continuous(labels = scales::percent)
```

## Reordering graphs

```{r}
usefulness_by_platform %>%
	# reorder learning_platform by avg_usefulness
	mutate(learning_platform = fct_reorder(learning_platform, avg_usefulness)) %>%
	# reverse the order of learning_platform
	mutate(learning_platform = fct_rev(learning_platform)) %>%
	ggplot(aes(x = learning_platform, y = avg_usefulness)) + 
    geom_point() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    labs(x = "Learning Platform", y = "Percent finding at least somewhat useful") + 
    scale_y_continuous(labels = scales::percent)
```

## Changing and creating variables with case_when()

### case_when()

```{r}
x <- 1:20
 
case_when(x %% 15 == 0 ~ "fizz buzz", 
          x %% 3 == 0 ~ "fizz", 
          x %% 5 == 0 ~ "buzz", 
          TRUE ~ as.character(x) )
```

### Order matters

```{r}
case_when(x %% 3 == 0 ~ "fizz buzz", 
          x %% 5 == 0 ~ "buzz", 
          x %% 3 == 0 ~ "fuzzy buzz",
          TRUE ~ as.character(x) )
```

## case_when() with multiple variables

```{r}
moods %>%
  mutate(action = case_when(
  mood == "happy" & status == "know it" ~ "clap your hands",
   mood == "happy" & status == "do not know it" ~ "stomp your feet", 
   mood == "sad" ~ "look at puppies", 
   TRUE ~ "jump around")
```

## case_when() with single variable

```{r}
# Check the min age
min(multiple_choice_responses$Age)
# Check the max age
max(multiple_choice_responses$Age)

multiple_choice_responses %>%
    # Filter for rows where Age is between 10 and 90
    filter(between(Age, 10,90)) %>%
    # Create the generation variable based on age
    mutate(generation = case_when(
    between(Age, 10, 22) ~ "Gen Z", 
    between(Age, 23, 37) ~ "Gen Y", 
    between(Age, 38, 52) ~ "Gen X", 
    between(Age, 53, 71) ~ "Baby Boomer", 
    between(Age, 72, 90) ~ "Silent"
    )) %>%
    # Get a count of how many answers in each generation
    count(generation)
```

## case_when() from multiple columns

```{r}
multiple_choice_responses %>%
    # Filter out people who selected Data Scientist as their Job Title
    filter(CurrentJobTitleSelect != "Data Scientist") %>%
    # Create a new variable, job_identity
    mutate(job_identity = case_when(
        CurrentJobTitleSelect == "Data Analyst" & 
        DataScienceIdentitySelect == "Yes" ~ "DS analysts", 
        CurrentJobTitleSelect == "Data Analyst" & 
        DataScienceIdentitySelect %in% c("No", "Sort of (Explain more)") ~ "NDS analyst", 
        CurrentJobTitleSelect != "Data Analyst" & 
        DataScienceIdentitySelect == "Yes" ~ "DS non-analysts", 
        TRUE ~ "NDS non analysts")) %>%
    # Get the average job satisfaction by job_identity, removing NAs
    group_by(job_identity) %>%
    summarise(avg_js = mean(JobSatisfaction, na.rm = T))
```

## Changing characters to factors

```{r}
flying_etiquette %>%
    # Change characters to factors
    mutate_if(is.character, as.factor) %>%
    # Filter out those who have never flown on a plane
    filter (`How often do you travel by plane?` != "Never")
```

## Tidying data

```{r}
gathered_data <- flying_etiquette %>%
    mutate_if(is.character, as.factor) %>%
    filter(`How often do you travel by plane?` != "Never") %>%
    # Select columns containing "rude"
    select(contains("rude")) %>%
    # Change format from wide to long
    gather("response_var", "value")
```

## Regex

```{r}
str_detec("happy", ".")
string <- "Statistics is the best"
str_remove(string, ".*the")
```

## Cleaning up strings

```{r}
gathered_data %>%
    # Remove everything before and including "rude to " (with that space at the end!)
    mutate(response_var = str_remove(response_var, '.*rude to ')) %>%
    # Remove "on a plane"
    mutate(response_var = str_remove(response_var, 'on a plane'))
```

## Dichotomizing variables

```{r}
dichotimized_data <- gathered_data %>%
    mutate(response_var = str_replace(response_var, '.*rude to ', '')) %>%
    mutate(response_var = str_replace(response_var, 'on a plane', '')) %>%
	# Remove rows that are NA in the value column
	filter(!is.na(value)) %>%
    # Dichotomize the value variable to make a new variable, rude
    mutate(rude = if_else(value %in% c('No, not rude at all', 'No, not at all rude'), 0, 1))
```

## Summarizing data

```{r}
rude_behaviors <- gathered_data %>%
    mutate(response_var = str_replace(response_var, '.*rude to ', '')) %>%
    mutate(response_var = str_replace(response_var, 'on a plane', '')) %>%
	# Remove rows that are NA in the value column
	filter(!is.na(value)) %>%
    mutate(rude = if_else(value %in% c("No, not rude at all", "No, not at all rude"), 0, 1)) %>%
    # Group by response_var
    group_by(response_var) %>%
    # Create perc_rude, the percent considering each behavior rude
    summarise(perc_rude = mean(rude))
```

## Recreating the plot

```{r}
ggplot(mtcars, aes(disp, mpg)) + geom_point() + 
  labs(x = "x axis label", y = "y axis label", title = "My title",
       subtitle = "and a subtitle", caption = "even a caption!")

initial_plot + 
  geom_text(aes(label = round(mean_mpg)))

initial_plot + 
  geom_text(aes(label = round(mean_mpg), 
                y = mean_mpg + 2))

initial_plot + 
  geom_text(aes(label = round(mean_mpg), y = mean_mpg + 2)) + 
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank())
```

## Creating an initial plot

```{r}
initial_plot <- rude_behaviors %>%
	# reorder response_var by perc_rude
	mutate(response_var = fct_reorder(response_var, perc_rude)) %>%
	# make a bar plot of perc_rude by response_var
	ggplot(aes(x = response_var, y = perc_rude)) + 
    geom_col() 

# View your plot
initial_plot
```

## Fixing labels

```{r}
titled_plot <- initial_plot + 
    # Add the title, subtitle, and caption
    labs(title = "Hell Is Other People In A Pressurized Metal Tube",
         subtitle = "Percentage of 874 air-passenger respondents who said action is very or somewhat rude",
         caption = "Source: SurveyMonkey Audience", 
         # Remove the x- and y-axis labels
         x = "", 
         y = "") 
```

## Flipping things around 

element_blank() is a nice way to keep your graphs looking tidy when you don't need the tick marks.

```{r}
flipped_plot <- titled_plot + 
    # Flip the axes
    coord_flip() + 
    # Remove the x-axis ticks and labels
    theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank())
```


## Finalizing the chart

```{r}
flipped_plot + 
    # Apply percent() to perc_rude to label above the bar with the perc value
    geom_text(aes(label = percent(perc_rude), 
                  y = perc_rude + .03), 
              position = position_dodge(0.9),
              vjust = 1)
```


