---
title: "Data Manipulation in R with data.table"
author: "Camila Bezerra"
date: "14/09/2019"
output:
  html_document:
    theme: united
    highlight: monochrome
---

```{r, warning=F}
library(ggplot2)
library(tidyverse)
library(dplyr)
```

# 1. Introduction to data.table

## data.table top quiz

* A data.table is a data.frame. It inherits and builds on top of it.
* data.table s never use or need row names

```{r}

```


## Creatinga data.table

```{r}
# Load data.table
library(data.table)

# Create the data.table X 
X <- data.table(is = c("a", "b", "c"), value = c(0.5, 1.0, 1.5))

# View X
X
```

##Introducing bikes data

```{r}
# Get number of columns in batrips
col_number <- ncol(batrips)

# Print the first 8 rows
head(batrips, n = 8)

# Print the last 8 rows
tail(batrips, n = 8)

 # Print the structure of batrips
str(batrips)

```

## Filtering rows using positive integers

```{r}
# Filter third row
row_3 <- batrips[3,]
row_3

# Filter rows 10 through 20
rows_10_20 <- batrips[10:20, ]
rows_10_20

# Filter the 1st, 6th and 10th rows
rows_1_6_10 <- batrips[c(1, 6, 10)]
rows_1_6_10
```

## Filtering rows using negative integers



## Filtering rows using negative integers

```{r}
# Select all rows except the first two
not_first_two <- batrips[-(1:2)]

# Select all rows except 1 through 5 and 10 through 15
exclude_some <- batrips[-c(1:5, 10:15)]
exclude_some

# Select all rows except the first and last
not_first_last <- batrips[!c(1, 326339)]
not_first_last

```

## Filtering rows using logical vectors

```{r}
# Filter all rows where start_station is "MLK Library"
trips_mlk <- batrips[start_station == "MLK Library"]
trips_mlk

# Filter all rows where start_station is "MLK Library" AND duration > 1600
trips_mlk_1600 <- batrips[start_station == "MLK Library" & duration > 1600]
trips_mlk_1600

# Filter all rows where `subscription_type` is not `"Subscriber"`
customers <- batrips[subscription_type != "Subscriber"]
customers

# Filter all rows where start_station is "Ryland Park" AND subscription_type is not "Customer"
ryland_park_subscribers <- batrips[start_station == "Ryland Park" & subscription_type != "Customer"]
ryland_park_subscribers
```

## Helpers for filtering

* %like% alllows you to search for a pattern in character or a factor vector
      * Usage: col %like% pattern in a character or a factor vector
      
```{r}
# SUbset all rows where start_station starts with San Francisco
batrips[start_station %like% "^San Francisco"]

# Instead of
batrips[grepl("^San Franciso", start_station)]
```

* %between5=%: allows you to search for values in the closed interval[val1, val2]
      * Usage: numeric_col %between% c(val1, val2)
      
```{r}
# subset all rows where duration is between 2000 and 3000
batrips[duration %between% c(2000, 3000)]

# Instead of
batrips[duration >= 2000 & duration <= 3000]
```

* %chin% is similar to %in%, but is much faster and oly for charcters
      * Usage: character_col %chin% c("val1", "val2", "val3")
      
```{r}
# Subset all rows  where start_station is
# "Japantown", "Mezes Park" or "MLK Library"
batrips[start_station %chin% c("Japantown", "Mezes Park", "MLK Library")]

# Much faster than
batrips[start_station %in% c("Japantown", "Mezes Park", "MLK Library")]
```

## I %like% data.tables

* Use %like% to filter all rows where end_station ends the text "Market"

```{r}
# Filter all rows where end_station ends with "Market" 
end_markets <- batrips[end_station %like% "Market$"]
## The metacharacter $ indicates that you are searching for the string that ends with a specific pattern.
```

## Filtering with %in%

```{r}
# Filter all rows where trip_id is 588841, 139560, or 139562
filter_trip_ids <- batrips[trip_id %in% c(588841, 139560, 139562)]
```

## Filtering with %between% and %chin%

```{r}
# Filter all rows where duration is between [5000, 6000]
duration_5k_6k <- batrips[duration %between% c (5000, 6000)]
duration_5k_6k

# Filter all rows with specific start stations
two_stations <- batrips[start_station %chin% c ("San Francisco City Hall", "Embarcadero at Sansome")]
two_stations
```

# 2. Selecting and Computing on Columns

## Selecting columns the data.table way

If you don't wrap yhe column name either in quotes or inside list(), the result is a vector.

Which of the following options return a data.table?

* batrips[, duration]

Which of the following return a data.table?
* batrips[, "duration"]
* batrips[, .(duration)]
* batrips[, list(duration)]

You can use a list of variables (column names) to select columns

```{r}
ans <- batrips[, list(trip_id, dur = duration)]
head(ans, 2)
```

* When selecting a single column, not wraping the variable by list() returns a vector

```{r}
# Select a single column and return a data.table
ans <- batrips[, list(trip_id)]

# Select a single column and return a vector
ans <- batrips{, trip_id}
```

* () is an allias to lits(), for convenience

```{r}
# .() is the same as list()
ans <- batrips[, .(trip_id, duration)]
head(ans, 2)
```

## Selecting columns by name

```{r}
# Select bike_id and trip_id using a character vector
df_way <- batrips[, c("bike_id", "trip_id")]

# Select start_station and end_station cols without a character vector
dt_way <- batrips[ , .(start_station, end_station)]
```

## Deselcting specific columns

```{r}
# Deselect start_terminal and end_terminal columns
drop_terminal_cols <- batrips[, -c("start_terminal", "end_terminal")]
drop_terminal_cols
```

## Computing on columns the data.table way

* since columns can be referred to as variables, you can compute directly on the in j

```{r}
# Compute mean o duration column using the data.table way
ans <- batrips[, mean(duration)]

# Compute mean of duration using the data.frame way
ans <- mean (batrips[, "duration"])
```

* Combining i and j is straightforward

```{r}
# Compute mean of duration column for "Japantown" start station 
batrips[start_station == "Japantown", mean(duration)]
```

* Special symbol .N in j

        * .N can be used in j as well
        * Particularly useful to get the number of rows after filtering in i

```{r}
# How many trips started from "Japantown"?
batrips[start_station == "Japantown", .N]

# Compare this to the data.frame way
nrow(batrips[batrips$start_station == "Japantown", ])
```

## Computing in j(I)

The data.table first filters the rows in i and then computes in j.

```{r}
# Calculate median duration using the j argument
median_duration <- median(batrips[, duration])

# Get median duration after filtering
median_duration_filter <- batrips[end_station == "Market at 10th" & subscription_type == "Subscriber", median(duration)]
median_duration_filter
```


## Computing in j(II)

```{r}
# Compute duration of all trips
trip_duration <- batrips[, difftime(end_date, start_date, units = "min")]
head(trip_duration)
```

## Advanced computation in j

Together with i, you can compute on columns in j only for those rows that satisfy a condition

```{r}
batrips[start_station == "Japantown", . (mn_dur = mean(duration),
med_dur = median(duration))]
```

## Computing in j(III)

```{r}
# Calculate the average duration as a one-column data.table with the column name mean_durn.
mean_duration <- batrips[, list(mean(duration))]

# Calculate the minimum and maximum duration (don't rename the columns).
min_max_duration <- batrips[, .(min(duration), max(duration))]
min_max_duration

# Find the average duration and the date of the last ride (using the end_date column). Name these columns mean_duration and last_ride, respectively.
# Calculate the number of unique values
other_stats <- batrips[, .(mean_duration= mean(duration), last_ride = max(end_date))]
other_stats
```

## Combining i and j

```{r}
# Calculate the minimum and maximum duration for all trips where the start_station is "Townsend at 7th" and the duration is less than 500. Rename these columns to min_dur and max_dur, respectively.

duration_stats <- batrips[start_station == "Townsend at 7th" & duration < 500,
                          .(min_dur = min(duration), 
                           max_dur = max(duration))]
duration_stats

# Plot the histogram of duration based on conditions
batrips[start_station == "Townsend at 7th" & duration < 500, hist(duration)]

```

# 3. Groupwise Operations

## Computations by groups

### The by argument

* The by argument allows computations for each unique value of the (grouping) columns specified in by

```{r}
# How many trips happened from each start_station?
ans <- batripts[, .N, by = "start_station"]
head(ans, 3)
```

* by argument accpets both character vector of column names as well as a list of variables/expressions

```{r}
# Same as batrips[, .N, by = "start_station"]
ans <- batrips[, .N, by = .(start_station)]
head(ans, 3)
```

* Allows renaming grouping columns on the fly

```{r}
ans <- batrips[, .(no_trips = .N), by = .(start = start_station)]
head(ans, 3)
```

* Expressions in by

The list() or .() expression in by allows for grouping variables to be computed on the fly

```{r}
# Get the number of trips for each start_station for each month
ans <- batrips[ , .N, by = .(start_station, mon = month(start_date))]
head(ans, 3)
```

## Computing stats by groups (I)

```{r}
# Compute the mean duration for every start_station
mean_start_stn <- batrips[, .(mean(duration)), by = start_station]
mean_start_stn
```

## Computing stats by groups (II)

```{r}
# Compute the mean duration for every start and end station
mean_station <- batrips[, .(mean_duration = mean(duration)), by = .(start_station, end_station)]
mean_station

# Compute the mean duration grouped by start_station and month
mean_start_station <- batrips[, .(mean_duration = mean(duration)), by = .(start_station, month(start_date))]
mean_start_station
```

## Computing multiple stats

Group by start_station and end_station to calculate the mean duration and the total number of trips. Name these columns mean_duration and total_trips, respectively.

```{r}
# Compute mean of duration and total trips grouped by start and end stations
aggregate_mean_trips <- batrips[, .(mean_duration = mean(duration), total_trips = .N), by = .(start_station, end_station)]
aggregate_mean_trips
```

Group by start_station, end_station, and for each month in start_date to calculate the minimum and maximum duration. Name these columns min_duration and max_duration, respectively.

```{r}
# Compute min and max duration grouped by start station, end station, and month
aggregate_min_max <- batrips[, .(min_duration = min(duration), max_duration = max(duration)), by = .(start_station, end_station, month(start_date))]
aggregate_min_max
```

## Chaining data.table expressions

```{r}
# Three start stations with the lowest mean duration
step_1 <- batrips[, .(mn_dur = mean(duration)), by = "start_station"]
step2 <- step_1(mn_dur)
step2[1:3]

# Three start stations with the lowest mean duration
batrips[, .(mn_dur = mean(duration)),
        by = "start_station"][order(mn_dur)][1:3]
```

### uniqueN()

* uniqueN() is a helper function that returns an integer value containing the number of unique values in the input object
* It accepts vectors as well as data.frames and data.tables

```{r}
id <- c(1, 2, 2, 1)
uniqueN(id)
```

### uniqueN() together with by

Calculate the total number of unique bike ids for every month

```{r}
ans <- batrips[, uniqueN(bike_id), by = month(start_date)]
head(ans, 3)
```

## Ordering rows

```{r}
# Arrange the total trips grouped by start_station and end_station in decreasing order
trips_dec <- batrips[, .N, by = .(start_station, 
                                  end_station)][order(-N)]
trips_dec
```

## What are the top 5 destinations?

```{r}
# Top five most popular destinations
top_5 <- batrips[, .N, by = end_station][order(-N)][1:5]
top_5
```

## What is the most popular destination from each start station?

```{r}
# Compute most popular end station for every start station
popular_end_station <- trips_dec[, .(end_station = end_station[1]), 
                                 by = start_station]
popular_end_station
```

## Combining i, j, and by (I)

```{r}
# Find the first and last ride for each start_station
first_last <- batrips[order(start_date),
                      .(start_date = start_date[c(1, .N)]),
                      by = .(start_station)]
```

## Computations in j using .SD

* .SD is a special symbol which stands for Subset of Data
* Contains subset of data corresponding to each group; which itself is a data table
* By default, the goruping columns are excluded for convenience

### Subset of Data, .SD

```{r}
x[, print(.SD), by = id]
```

* SDcols holds the columns that should be included in .SD

```{r}
# .SDcols controls the columns .SD contains
batrips[, .SD[1], by = start_station, .SDcols = c("trip_id", "duration")]

batrips[, .SD[1], by = start_station, .SDcols = -c("trip_id", "duration")]
```

## Using.SD(I)

```{r}
relevant_cols <- c("start_station", "end_station", 
                   "start_date", "end_date", "duration")

# Find the row corresponding to the shortest trip per month
shortest <- batrips[, .SD[which.min(duration)], 
                    by = month(start_date), 
                    .SDcols = relevant_cols]
shortest
```

## Using .SD (II)

```{r}
# Find the total number of unique start stations and zip codes per month
unique_station_month <- batrips[, lapply(.SD, uniqueN),
                                by = month(start_date),
                                .SDcols = c("start_station", "zip_code")]
unique_station_month
```

# 4. Reference Semantics

## data.table internals

* data.table updates columns in place, i.e., by reference
* This means, you don't need the assign the result back to a variable
* No copy of any column is made while their values are changed
* data.table uses a new opeartor:= to add/update/delete columns by reference

### LHS := RHS form

```{r}
batrips[, c ("is_dur_gt_1hour", "week_day") := list(duration > 3600, wday(start_date))]

# When adding a single column quotes aren't necessary 
batrips[, is_dur_gt_1hour := duration > 3600]
```

### Functional form

```{r}
batrips[, `=`(is_dur_gt_1hiur = NULL, start_station = toupper(start_station))]
```

## Adding a new column

```{r}
# Add a new column, duration_hour
batrips[, duration_hour := duration/3600]
```

## Updating an existing column (I) 

```{r}
# Print untidy
head(untidy, n = 2)

# Fix spelling in the second row of start_station
untidy[2, start_station := "San Francisco City Hall"]
```

## Updating an existing column(II)

```{r}
# Replace negative duration values with NA
untidy[duration <0, duration := NA]
```

## Combining ":=" with by

```{r}
batrips[, n_zip_code := .N, by = zip_code]

zip_1000 <- batrips[n-zip_code > 1000][, n_zip_code := NULL]

# Same as 
zip_1000 <- bartrips[, n_zip_code := .N,
                     by = zip_code][n_zip_code > 1000][, n_zip_code := NULL]
```

## Adding columns by group

```{r}
# Add a new column equal to total trips for every start station
batrips[, trips_N := .N, by = start_station]

# Add new column for every start_station and end_station
batrips[, duration_mean := mean(duration), by = c("start_station", "end_station")]
```

## Updating columns by group

Add a new column (mean_dur) by reference that is the mean duration of all trips grouped by month (based on start_date). Note that the duration column has missing values.

```{r}
# Calculate the mean duration for each month
batrips_new[, mean_dur := mean(duration, na.rm = TRUE), 
            by = month(start_date)]
```

Chain a new data.table expression that replaces all missing values in duration with the corresponding mean_dur value.


```{r}
# Replace NA values in duration with the mean value of duration for that month
batrips_new[, mean_dur := mean(duration, nna.rm = TRUE),
            by = month(start_date)][is.na(duration),
                                    duration := mean_dur]

# Delete the mean_dur column by reference
batrips_new[, mean_dur := mean(duration, na.rm = TRUE), 
            by = month(start_date)][is.na(duration), 
                                    duration := mean_dur][, mean_dur := NULL]
```

## Advanced multiple columns by reference by group

```{r}
# Functional form
batrips[, `:=`(end_dur_first = duration[1],
               end_dur_last = duration[.N]),
        by = end_station]

# LHS := RHS form
batrips[, c("end_dur_first",
            "end_dur_last") := list(duration[1]), duration[.N]),
by = end_station]
```


## Adding multiple columns (I)

```{r}
# Add columns using the LHS := RHS form
batrips[, c("mean_duration", "median_duration") := list(mean(duration), median(duration)), 
        by = start_station]
```

## Adding multiple columns (II)

```{r}
# Add columns using the functional form
batrips[,`:=`(mean_duration = mean(duration),
median_duration = median(duration)), 
        by = start_station]
```

## Combining i, j, and by (II)

```{r}
# Add the mean_duration column
batrips[duration > 600, mean_duration := mean(duration), by = .(start_station, end_station)]
```

# 5. Importing and Exporting Data

## Fast data reading with fread()

```{r}
# file from URL
DT1 <- fread("https://bit.ly/uojpwç")
DT1

# Local file
DT2 <- fread("data.csv")
DT2

# String
DT3 <- fread("a,b\n1,2\n3,4")
DT3

# String without col names
DT4 <- fread("1,2\n3,4")
DT4
```

### nrows and skip arguments

```{r}
# Read only first line(after header)
fread("a,b\n1,2\n3,4", nrows = 1)

# Skip first two lines containing metadata
str <- "#Metadata\nTimestamp: 2018-05-01 19:44:28 GMT\na,b\n1,2\n3,4"
fread(str, skip = 2)

str <- "#Metadata\nTimestamp: 2018-05-01 19:44:28 GMT\na,b\n1,2\n3,4"
fread(str, skip = "a,b")

fread(str, skip = "a,b", nrows = 1)
```

### select and drop arguments

```{r}
str <- "a,b,c\n1,2,x\n3,4,y"
fread(str, select = c("a", "c"))

# Same as 
fread(str, drop = "b")

str <- "1,2,x\n3,4,y"
fread(str, select = c(1,3))

# Same as
fread(str, drop = 2)
```

## Fast reading from disk

```{r}
# Use read.csv() to import batrips
system.time(read.csv("batrips.csv"))

# Use fread() to import batrips
system.time(fread("batrips.csv"))
```


## Importing a CSV file

fread( ) is not only fast but also makes things a lot easier by guessing sensibly.
Unlike the additional arguments which we provided for read.csv(), fread() could handle importing the file by itself.
Also note the useful warning fread() provides.

```{r}
# Import using read.csv()
csv_file <- read.csv("sample.csv", fill = NA, quote = "", 
                     stringsAsFactors = FALSE, strip.white = TRUE, 
                     header = TRUE)
csv_file

# Import using fread()
csv_file <- fread("sample.csv")
csv_file
```


## Importing selected columns

```{r}
# Select "id" and "val" columns
select_columns <- fread("sample.csv", select = c("id", "val"))
select_columns

# Drop the "val" column
drop_column <- fread("sample.csv", drop = "val")
drop_column
```

## Importing selected rows

```{r}
# Import the file
entire_file <- fread("sample.csv")
entire_file

# Import the file while avoiding the warning
only_data <- fread("sample.csv", nrows = 4)
only_data

# Import only the metadata
only_metadata <- fread("sample.csv", skip = 7)
only_metadata
```


## Advanced file reading

* By default, R can only represent numbers less than or equal to 2^31-1 = 2147...
* Large integers are autimatically read as in integer64 type, provided by tge bit64 package

```{r}
ans <- fread("id,name\n1234567890123,Jane\n5284782381811,John\n")
class(ans$id)
#"integer64"
```

### Specifying column class types with colClasses

```{r}
ans <- fread(str, colClasses = c("integer", "integer", "numeric", "logical", "factor"))
```


### The fill argument

```{r}
fread(str, fill = TRUE)
```

## Reading large integers

Well done! Whenever possible, you should import your data as integers, especially when your data is big. Doing so results in saving memory in your RAM.

```{r}
# Import the file using fread 
fread_import <- fread("sample.csv")

# Import the file using read.csv 
base_import <- read.csv("sample.csv")

# Check the class of id column
class(fread_import$id)
class(base_import$id)
```

## Specifying column classes

```{r}
# Import using read.csv with defaults
base_r_defaults <- read.csv("sample.csv")
str(base_r_defaults)

# Import using read.csv
base_r <- read.csv("sample.csv", 
                   colClasses = c(rep("factor", 4), 
                                  "character", "integer", 
                                  rep("numeric", 4)))
str(base_r)

# Import using fread
import_fread <- fread("sample.csv", 
                      colClasses = list(factor = 1:4, numeric = 7:10))

str(import_fread)
```

##

* fread() considered the last line as a footer and discarded it. Import the file (sample.csv) using fread() again, but this time, use the fill argument to ensure all rows are imported correctly.

```{r}
# Import the file correctly
correct <- fread("sample.csv", fill = TRUE)
```


## Dealing with missing values

```{r}
# Import the file using na.strings
missing_values <- fread("sample.csv",na.strings = c("##") )
```


## Fast data writing with fwrite()

Ability to write list columns using secondary operator(|)

```{r}
dt <- data.table(id = c("x", "y", "z"), val = list(1:2, 3:4, 5:6))
fwrite(dt, "fwrite.csv")
fread("fwrite.csv")
```

### data and datetime columns(ISO)

* fwrite() provides three aditional ways of writing date and datetime format - ISO, squash and epoch
* Encourages the use of ISO standards with ISO as default

```{r}
now <- Sys.time()
dt <- data.table(date = as.IDate(now),
                 time = as.ITime(now),
                 datetime = now)

# ISO is default
fwrite(dt, "datetime.csv", dateTimeAs = "ISO")
fread("datetime.csv")
```


## Writing files to disk

```{r}
# Write dt to fwrite.txt
fwrite(dt, "fwrite.txt")

# Import the file using readLines()
readLines("fwrite.txt")

# Import the file using fread()
fread("fwrite.txt")
```

## Writing date and time columns

```{r}
# Write batrips_dates to file using "ISO" format
fwrite(batrips_dates, "iso.txt",dateTimeAs = "ISO")

# Import the file back
iso <- fread("iso.txt")
iso

# Write batrips_dates to file using "squash" format
fwrite(batrips_dates, "squash.txt", dateTimeAs = "squash")

# Import the file back
squash <- fread("squash.txt")
squash

# Write batrips_dates to file using "epoch" format
fwrite(batrips_dates, "epoch.txt", dateTimeAs = "epoch")

# Import the file back
epoch <- fread("epoch.txt")
epoch
```

## Fast writing to disk

```{r}
# Use write.table() to write batrips
system.time(write.table(batrips, "base-r.txt"))

# Use fwrite() to write batrips
system.time(fwrite(batrips, "data-table.txt"))
```

