---
title: "Importing & Cleaning Data in R Case Studies"
author: "Camila Bezerra"
date: "16/08/2019"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```

# 1. Ticket Sales Data

```{r}
# Import sales.csv: sales
sales <- read.csv("sales.csv", stringsAsFactors = FALSE)

# View dimensions of sales
dim(sales)

# Inspect first 6 rows of sales
head(sales)

# View column names of sales
names(sales)

# Look at structure of sales
str(sales)

# View a summary of sales
summary(sales)

# Load dplyr
library(dplyr)

# Get a glimpse of sales
glimpse(sales)
```

## Removing redundant info

You can use square brackets to subset data frame:

```{r}
my_dif[1:5, ] # First 5 rows of my_df
my_df[, 4] #Fourth column of my_df
```


You can remove rows and columns using negative indices.

```{r}
my_dif[-(1:5), ] # Omit first 5 rows of my_df
my_dif[ , -4] # Omit fourth column of my_df
```


```{r}
# Remove the first column of sales: sales2
sales2 <- sales[, -1]

#Define a vector of column indices: keep
keep <- seq(5, ncol(sales2)-15, 1)

# Subset sales2 using keep: sales3
sales3 <- sales2[ ,keep]
```


```{r}
# Load tidyr
library(tidyr)

# Split event_date_time: sales4
sales4 <- separate(sales3, event_date_time,
                   c("event_dt", "event_time"), sep = " ")

# Split sales_ord_create_dttm: sales5
sales5 <- separate(sales4, sales_ord_create_dttm, c("ord_create_dt", "ord_create_time"), sep = " ")
```

## Identifying dates

```{r}
#Load stringr
library(stringr)

#Find columns of sale5 containing "dt":date_cols
date_cols <- str_detect(colnames(sales5), "dt")

#Load lubridate
library(lubridate)

#Coerce date columns into Date objects
sales5[, date_cols] <- lapply(sales5[, date_cols], ymd)
```

## More warnings!

```{r}
#Create a numerical vector that counts missing values: num_missing
num_missing <- sapply(missing, sum)

#Print num_missing
num_missing
```

## Combining columns
```{r}
# Combine the venue_city and venue_state columns
sales6 <- unite(sales5, venue_city_state, venue_city, venue_state, sep = " ")

#View the head of sales6
head(sales6)
```


# 2. MBTA Ridership Data

## Using readxl
```{r}
#Load readxl
library(readxl)
```

## Removing unnecessary rows and columns 

```{r}
# Remove rows 1, 7, and 11 of mbta: mbta2
mbta2 <- mbta[-c(1, 7, 11), ]

# Remove the first column of mbta2: mbta3
mbta3 <- mbta2[, -1]
```

## Observations are stored in columns
```{r}
#Load tidyr
library(tidyr)

# Gather columns of mbta3: mbta4
mbta4 <- gather(mbta3, month, thou_riders, -mode)

#View the head of mbta4
head(mbta4)

# Coerce thou_riders to numeric
mbta4$thou_riders <- as.numeric(mbta4$thou_riders)
```

## Variables are stored in both rows and columns

```{r}
# Spread the contents of mbta4: mbta5
mbta5 <- spread(mbta4, mode, thou_riders)

# View the head of mbta5
head(mbta5)
```

##Separating columns

```{r}
# View the head of mbta5
head(mbta5)

# Split month column into month and year: mbta6
mbta6 <- separate(mbta5, month, c('year', 'month'))

# View the head of mbta6
head(mbta6)
```

## Dealing with entry error
```{r}
# Find the row number of the incorrect value: i
i <- which(mbta6$Boat == 40)

# Replace the incorrect value with 4
mbta6$Boat[i] <- 4

# Generate a histogram of Boat column
hist(mbta6$Boat)

# Look at Boat and Trackless Trolley ridership over time (don't change)
ggplot(mbta_boat, aes(x = month, y = thou_riders, col = mode)) +  geom_point() + 
  scale_x_discrete(name = "Month", breaks = c(200701, 200801, 200901, 201001, 201101)) + 
  scale_y_continuous(name = "Avg Weekday Ridership (thousands)")

# Look at all T ridership over time (don't change)
ggplot(mbta_all, aes(x = month, y = thou_riders, col = mode)) + geom_point() + 
  scale_x_discrete(name = "Month", breaks = c(200701, 200801, 200901, 201001, 201101)) +  
  scale_y_continuous(name = "Avg Weekday Ridership (thousands)")
```


# 3. World Food Facts

## Importing the data

```{r}
# Load data.table
library(data.table)

# Import food.csv as a data frame: food
food <- fread("food.csv", data.table = F)
```

## Removing duplicate info

```{r}
my_dif[, -3] #Omit third column
```

```{r}
# Define vector of duplicate cols (don't change)
duplicates <- c(4, 6, 11, 13, 15, 17, 18, 20, 22, 
                24, 25, 28, 32, 34, 36, 38, 40, 
                44, 46, 48, 51, 54, 65, 158)

# Remove duplicates from food: food2
food2 <- food[, -duplicates]
```

##Removing useless info

```{r}
# Define useless vector (don't change)
useless <- c(1, 2, 3, 32:41)

# Remove useless columns from food2: food3
food3 <- food2[, -useless]
```

```{r}
# Create a vector of column indices: nutrition
nutrition <- str_detect(names(food3), "100g")

# View a summary of nutrition columns
summary(food3[, nutrition])
```


##Replacing missing values

```{r}
# Find indices of sugar NA values: missing
missing <- is.na(food3$sugars_100g)

# Replace NA values with 0
food3$sugars_100g[missing] <- 0

# Create first histogram
hist(food3$sugars_100, breaks = 100)

# Create food4
food4 <- food3[food3$sugars_100g != 0, ]

# Create second histogram
hist(food4$sugars_100g, breaks = 100)
```


## Dealing with messy data

```{r}
# Find entries containing "plasti": plastic
plastic <- str_detect(food3$packaging, "plasti")

# Print the sum of plastic
sum(plastic)
```


# 4. School Attendance

## Importing the data

```{r}
# Load the gdata package
library(gdata)

# Import the spreadsheet: att
att <- read.xls("attendance.xls")
```

## Examining the data

```{r}
# Print the column names 
names(att)

# Print the first 6 rows
head(att)

# Print the last 6 rows
tail(att)

# Print the structure
str(att)
```

## Removing uncessary rows

```{r}
# Create remove
remove <- c(3, 56, 57, 58, 59)

# Create att2
att2 <- att[-remove, ]
```

## Removing useless columns

```{r}
# Create remove
remove <- c(3, 5, 7, 9, 11, 13, 15, 17)

# Create att3
att3 <- att2[, -remove]
```

## Splitting the data

```{r}
# Subset just elementary schools: att_elem
att_elem <- att3[ , c(1, 6, 7)]

# Subset just secondary schools: att_sec
att_sec <- att3[ , c(1, 8, 9)]

# Subset all schools: att4
att4 <- att3[  , c(1:5)]
```

## Replacing the names

```{r}
# Define cnames vector (don't change)
cnames <- c("state", "avg_attend_pct", "avg_hr_per_day", 
            "avg_day_per_yr", "avg_hr_per_yr")

# Assign column names of att4
colnames(att4) <- cnames

# Remove first two rows of att4: att5
att5 <- att4[ -c(1,2), ]

# View the names of att5
names(att5)
```

## Cleaning up extra characters

```{r}
## Remove all periods in state column
att5$state <- str_replace_all(att5$state, "\\.", "")

## Remove white space around state names
att5$state <- str_trim(att5$state)

## View the head of att5
head(att5)
```

## Some final type conversions

```{r}
# Change columns to numeric using dplyr (don't change)
library(dplyr)
example <- mutate_at(att5, vars(-state), funs(as.numeric))

# Define vector containing numerical columns: cols
cols <- 2:ncol(att5)

# Use sapply to coerce cols to numeric
att5[, cols] <- sapply(att5[, cols], as.numeric)
```

#Some final type conversions

```{r}
# Change columns to numeric using dplyr (don't change)
library(dplyr)
example <- mutate_at(att5, vars(-state), funs(as.numeric))

# Define vector containing numerical columns: cols
cols <- -1

# Use sapply to coerce cols to numeric
att5[, cols] <- att5[, cols] <- sapply(att5[,cols], as.numeric)
```

